% Blueprint for "Fundamental Limitations on Subquadratic Alternatives to
% Transformers" (Josh Alman, Hantao Yu, ICLR 2025), arXiv:2410.04271v2.
%
% Two-kind model:
%   * LEAF nodes (\mathlibok, no proof) = standard facts already in Mathlib
%     (inner product, norm, continuity closure).
%   * FULL nodes = the paper's definitions, hypotheses, reductions, hardness
%     theorems and transformer constructions. These carry a complete statement,
%     \uses edges, and (for results) a complete proof. They are NOT marked
%     \leanok/\mathlibok.
% Node titles keep the paper's numbering, e.g. [Theorem 3.1], [Lemma A.8].

\chapter{Preliminaries: notation and library facts}

% --- Settled dependencies: already in Mathlib, emitted as leaves. -----------

\begin{definition}[Inner product of binary/real vectors]
  \label{def:inner-prod}
  \lean{inner}
  \mathlibok
  For $v, w \in \R^{\ell}$ we write $\inner{v}{w} = \sum_{i=1}^{\ell} v[i]\, w[i]$
  for their (standard) inner product, where $v[i]$ denotes the $i$-th entry of
  $v$. Equivalently, for vectors indexed by a finite type $I$, one may write
  $\inner{v}{w} = \sum_{i \in I} v_i w_i$. For binary vectors
  $v,w \in \{0,1\}^{\ell}$ this counts the coordinates on which both are $1$.
  Recalled only so later nodes can depend on it.
\end{definition}

\begin{definition}[$\ell_2$ norm]
  \label{def:l2norm}
  \lean{EuclideanSpace.norm_eq}
  \mathlibok
  For $v \in \R^{\ell}$, $\norm{v} = \sqrt{\inner{v}{v}}$ is the $\ell_2$ norm.
  We also use the $\ell_1$ norm $\norm{v}_1 = \sum_{i=1}^{\ell} |v[i]|$; for a
  binary vector $\norm{v}_1$ is its number of $1$ entries. Standard.
\end{definition}

\begin{definition}[Kronecker (tensor) product of vectors]
  \label{def:kron}
  For finite index sets $I,J$, and vectors $v : I \to \R$ and $w : J \to \R$,
  the Kronecker product $v \otimes w : I \times J \to \R$ is the vector defined
  entrywise by
  \[
    (v \otimes w)_{(i,j)} = v_i w_j .
  \]
  For vectors in $\R^a$ and $\R^b$ this is the same vector in $\R^{ab}$ whose
  entries are all products of one entry of $v$ and one entry of $w$. We write
  $v^{\otimes q}$ for the $q$-fold Kronecker power $v \otimes \cdots \otimes v$.
  This is the vector-indexed Kronecker product used in the paper.
\end{definition}

\begin{lemma}[Finite sum over a product type]
  \label{lem:sum-prod-type}
  \lean{Fintype.sum_prod_type}
  \mathlibok
  For finite types $I,J$ and a function $h : I\times J \to \R$,
  \[
    \sum_{p\in I\times J} h(p)
    =
    \sum_{i\in I}\sum_{j\in J} h(i,j).
  \]
  This is the uncurried product-index expansion. In Lean, this is the
  unprimed theorem `Fintype.sum_prod_type`; the primed theorem is for a curried
  function and should not be used when the target is already a sum over
  $p : I\times J$.
\end{lemma}

\begin{lemma}[Product of two finite sums]
  \label{lem:sum-mul-sum}
  \lean{Fintype.sum_mul_sum}
  \mathlibok
  For finite types $I,J$ and functions $f : I\to\R$ and $g : J\to\R$,
  \[
    \left(\sum_{i\in I} f(i)\right)
    \left(\sum_{j\in J} g(j)\right)
    =
    \sum_{i\in I}\sum_{j\in J} f(i)g(j).
  \]
  Equivalently, the displayed double sum factors as the product of the two
  finite sums by using `Fintype.sum_mul_sum` in the reverse direction.
\end{lemma}

\begin{lemma}[Multiplicativity of inner products under Kronecker product]
  \label{lem:kron-inner}
  \uses{def:inner-prod, def:kron, def:l2norm, lem:sum-prod-type, lem:sum-mul-sum}
  For $v,w \in \R^{a}$ and $x,y\in\R^{b}$,
  \[
    \inner{v\otimes x}{w\otimes y}
    =
    \inner{v}{w}\cdot\inner{x}{y}.
  \]
  This node is only the binary Kronecker identity. The $q$-fold consequences
  used in Appendix B are separate nodes
  Lemma~\ref{lem:kron-power-inner}, Lemma~\ref{lem:kron-power-norm}, and
  Lemma~\ref{lem:kron-power-cosine}. A formalization should not bundle those
  conclusions into this theorem; doing so creates a proof obligation whose
  induction and indexing data are absent from the binary statement.
\end{lemma}
\begin{proof}
  \uses{def:inner-prod, def:kron, def:l2norm, lem:sum-prod-type, lem:sum-mul-sum}
  Expanding the definitions over finite index sets $I,J$ gives a sum indexed by
  the product type:
  \[
    \inner{v\otimes x}{w\otimes y}
    = \sum_{p\in I\times J} (v_{p_1} x_{p_2})(w_{p_1} y_{p_2}).
  \]
  First commute and reassociate the scalar factors in each summand:
  \[
    (v_{p_1}x_{p_2})(w_{p_1}y_{p_2})
    =
    (v_{p_1}w_{p_1})(x_{p_2}y_{p_2}).
  \]
  Then use Lemma~\ref{lem:sum-prod-type} with
  $h(i,j)=(v_iw_i)(x_jy_j)$ to rewrite the product-indexed sum as
  \[
    \sum_{i\in I}\sum_{j\in J} (v_i w_i)(x_j y_j).
  \]
  Finally apply Lemma~\ref{lem:sum-mul-sum} in the reverse direction, with
  $f(i)=v_iw_i$ and $g(j)=x_jy_j$, to factor this double sum as
  \[
    \left(\sum_{i\in I} v_i w_i\right)
    \left(\sum_{j\in J} x_j y_j\right)
    = \inner{v}{w}\inner{x}{y}.
  \]
  Iterating this identity proves the statement for $q$-fold Kronecker powers.
  Applying the same identity with $w=v$ and $y=x$, and then taking square roots,
  gives $\norm{v\otimes x}=\norm{v}\norm{x}$ for nonnegative Euclidean norms;
  iterating gives $\norm{v^{\otimes q}}=\norm{v}^{q}$.
\end{proof}

\begin{lemma}[Inner products of Kronecker powers]
  \label{lem:kron-power-inner}
  \uses{def:inner-prod, def:kron, lem:kron-inner}
  For every finite index set $I$, every $v,w:I\to\R$, and every natural
  $q$, where $v^{\otimes q}$ and $w^{\otimes q}$ are indexed by functions
  $\{1,\ldots,q\}\to I$,
  \[
    \inner{v^{\otimes q}}{w^{\otimes q}}=\inner{v}{w}^{q}.
  \]
  The formal statement must specify the indexing convention for the $q$-fold
  power, for example the index type $\mathrm{Fin}\,q\to I$ with the empty product
  equal to $1$ when $q=0$.
\end{lemma}
\begin{proof}
  \uses{def:inner-prod, def:kron, lem:kron-inner}
  Induct on $q$.  For $q=0$, both sides are $1$: the Kronecker power has one
  coordinate indexed by the unique function from the empty index set, its entry
  is the empty product, and the scalar power $\inner{v}{w}^{0}$ is $1$.

  For the induction step, identify the index type for $q+1$ factors with
  $I\times(\mathrm{Fin}\,q\to I)$, separating the first tensor coordinate from
  the remaining $q$ coordinates. Under this identification
  \[
    v^{\otimes(q+1)} = v\otimes v^{\otimes q},
    \qquad
    w^{\otimes(q+1)} = w\otimes w^{\otimes q}.
  \]
  Lemma~\ref{lem:kron-inner} gives
  \[
    \inner{v^{\otimes(q+1)}}{w^{\otimes(q+1)}}
    =
    \inner{v}{w}\cdot
    \inner{v^{\otimes q}}{w^{\otimes q}}.
  \]
  Substitute the induction hypothesis and simplify scalar powers to obtain
  $\inner{v}{w}^{q+1}$.
\end{proof}

\begin{lemma}[Euclidean norms of Kronecker powers]
  \label{lem:kron-power-norm}
  \uses{def:l2norm, lem:kron-power-inner}
  For every finite index set $I$, every vector $v:I\to\R$, and every natural
  $q$,
  \[
    \norm{v^{\otimes q}}=\norm{v}^{q}.
  \]
  The proof must use nonnegativity of Euclidean norms and the square identity
  from Lemma~\ref{lem:kron-power-inner}; it is not a purely ring-theoretic
  consequence.
\end{lemma}
\begin{proof}
  \uses{def:l2norm, lem:kron-power-inner}
  Apply Lemma~\ref{lem:kron-power-inner} with $w=v$:
  \[
    \inner{v^{\otimes q}}{v^{\otimes q}}=\inner{v}{v}^{q}.
  \]
  By Definition~\ref{def:l2norm}, the left-hand side is
  $\norm{v^{\otimes q}}^{2}$ and $\inner{v}{v}=\norm{v}^{2}$. Hence
  \[
    \norm{v^{\otimes q}}^{2}=(\norm{v}^{2})^{q}=(\norm{v}^{q})^{2}.
  \]
  Both $\norm{v^{\otimes q}}$ and $\norm{v}^{q}$ are nonnegative, so equality of
  their squares implies equality of the norms.
\end{proof}

\begin{lemma}[Cosine similarity of Kronecker powers]
  \label{lem:kron-power-cosine}
  \uses{def:cosine, lem:kron-power-inner, lem:kron-power-norm}
  For nonzero vectors $v,w\in\R^{\ell}$ and every natural $q$,
  \[
    \cossim{v^{\otimes q}}{w^{\otimes q}}
    =
    \left(\cossim{v}{w}\right)^{q}.
  \]
  This is the exact identity needed in Lemma~\ref{lem:msd-tensor-reduction}.
\end{lemma}
\begin{proof}
  \uses{def:cosine, lem:kron-power-inner, lem:kron-power-norm}
  Expand cosine similarity and apply Lemma~\ref{lem:kron-power-inner} to the
  numerator and Lemma~\ref{lem:kron-power-norm} to both denominator factors:
  \[
    \cossim{v^{\otimes q}}{w^{\otimes q}}
    =
    \frac{\inner{v}{w}^{q}}{\norm{v}^{q}\norm{w}^{q}}.
  \]
  Since $v$ and $w$ are nonzero, their norms are nonzero. Rearranging powers in
  the field gives
  \[
    \frac{\inner{v}{w}^{q}}{\norm{v}^{q}\norm{w}^{q}}
    =
    \left(\frac{\inner{v}{w}}{\norm{v}\norm{w}}\right)^q
    =
    \left(\cossim{v}{w}\right)^q.
  \]
\end{proof}

\begin{lemma}[Continuity is closed under sums, products and composition]
  \label{lem:continuity-closure}
  \lean{Continuous.comp}
  \mathlibok
  This leaf packages standard continuity declarations such as
  `Continuous.add`, `Continuous.sub`, `Continuous.mul`, `Continuous.comp`,
  `continuous_const`, and continuity of coordinate projections. Finite sums,
  finite products and compositions of continuous functions between
  finite-dimensional real vector spaces are continuous, as are constant maps,
  coordinate projections and affine scalar maps. Recalled so the
  piecewise-linear MLP constructions below can be shown continuous.
\end{lemma}

\begin{lemma}[Continuity of max and min with a continuous argument]
  \label{lem:continuous-real-max-min}
  \lean{Continuous.max, Continuous.min}
  \mathlibok
  If $h : X \to \R$ is continuous and $c\in\R$ is constant, then the maps
  $x\mapsto \max(c,h(x))$ and $x\mapsto \min(c,h(x))$ are continuous. In Lean
  these are not proved by treating `max c` or `min c` as independently
  continuous curried functions; apply
  `Continuous.max continuous_const hh` and
  `Continuous.min continuous_const hh` to the full lambda expression.
\end{lemma}

\begin{lemma}[Continuity of the affine divided map]
  \label{lem:continuous-affine-divided}
  \uses{lem:continuity-closure}
  For real numbers $a,b$ with $b>a$, the map
  \[
    y\mapsto \frac{y-a}{b-a}
  \]
  is continuous as a map $\R\to\R$.
\end{lemma}
\begin{proof}
  \uses{lem:continuity-closure}
  Since $b>a$, the denominator $b-a$ is nonzero. Rewrite division by the
  constant $b-a$ as multiplication by the constant $(b-a)^{-1}$:
  \[
    \frac{y-a}{b-a}=(y-a)(b-a)^{-1}.
  \]
  The map $y\mapsto y-a$ is continuous by continuity of the identity map,
  constants and subtraction, and multiplication by the constant $(b-a)^{-1}$
  preserves continuity. Thus the displayed affine-divided map is continuous.

  A Lean proof should expose this as a separate continuous term, for example:
  first prove `hden : b - a \ne 0`; then prove
  `h_aff : Continuous (fun y : \R => (y - a) / (b - a))` by rewriting
  `div_eq_mul_inv` and using continuity of subtraction and multiplication by a
  constant. This proof should not be hidden inside a later `continuity` call.
\end{proof}

\begin{lemma}[Continuity of finite coordinate products]
  \label{lem:continuous-finite-product}
  \lean{continuous_finsetProd}
  \mathlibok
  Let $I$ be finite and let $h_i : X\to\R$ be continuous for every $i\in I$.
  Then $x\mapsto \prod_{i\in I} h_i(x)$ is continuous. In Lean, for a product
  over all coordinates $i:\mathrm{Fin}\,\ell$, use
  `continuous_finsetProd Finset.univ` and prove each factor continuous.
\end{lemma}

\begin{lemma}[Continuity of the clipped affine threshold]
  \label{lem:continuous-clipped-affine}
  \uses{lem:continuous-affine-divided, lem:continuous-real-max-min}
  For real numbers $a,b$ with $b>a$, the function
  \[
    g(y)=\max\!\left(0,\min\!\left(1,\frac{y-a}{b-a}\right)\right)
  \]
  is continuous as a map $\R\to\R$.
\end{lemma}
\begin{proof}
  \uses{lem:continuous-affine-divided, lem:continuous-real-max-min}
  By Lemma~\ref{lem:continuous-affine-divided}, the inner affine-divided map
  \[
    h(y)=\frac{y-a}{b-a}
  \]
  is continuous. Apply Lemma~\ref{lem:continuous-real-max-min} to the continuous
  map $h$ and the constant $1$ to obtain continuity of
  \[
    y\mapsto \min(1,h(y)).
  \]
  Call this continuous map $h_{\min}$. Applying
  Lemma~\ref{lem:continuous-real-max-min} again, now to $h_{\min}$ and the
  constant $0$, gives continuity of
  \[
    y\mapsto \max(0,h_{\min}(y))
    =
    \max\!\left(0,\min\!\left(1,\frac{y-a}{b-a}\right)\right).
  \]

  For a Lean proof, do not use the tactic `continuity` for this node, because it
  may create unprovable curried goals such as `Continuous (max 0)` or
  `Continuous (min 1)`. The intended proof is the direct one:
  define
  `h_aff : Continuous (fun y : \R => (y - a) / (b - a))` using
  Lemma~\ref{lem:continuous-affine-divided}; then set
  `h_min : Continuous (fun y : \R => min 1 ((y - a) / (b - a))) :=
  Continuous.min continuous_const h_aff`; finally conclude with
  `Continuous.max continuous_const h_min`.
\end{proof}

\begin{lemma}[Values of the clipped affine threshold]
  \label{lem:clipped-affine-values}
  \uses{lem:continuous-clipped-affine}
  For real numbers $a,b$ with $b>a$, define
  \[
    g(y)=\max\!\left(0,\min\!\left(1,\frac{y-a}{b-a}\right)\right).
  \]
  Then
  \[
    y\ge b \Longrightarrow g(y)=1,
    \qquad
    y<a \Longrightarrow g(y)=0.
  \]
  The two endpoint behaviors are part of this node. Continuity alone is not
  enough for Lemma~\ref{lem:mlp-threshold}.
\end{lemma}
\begin{proof}
  \uses{lem:continuous-clipped-affine}
  Since $b>a$, the denominator $b-a$ is strictly positive. If $y\ge b$, then
  $y-a\ge b-a$, so $(y-a)/(b-a)\ge1$. Hence the inner minimum is $1$, and the
  outer maximum is also $1$.

  If $y<a$, then $y-a<0$, so $(y-a)/(b-a)<0$. Hence the inner minimum is at
  most this negative number, and the outer maximum with $0$ is $0$.
\end{proof}

\begin{lemma}[Finite products with one zero factor and all-one factors]
  \label{lem:finite-product-zero-one}
  \lean{Finset.prod_eq_zero, Finset.prod_congr}
  \mathlibok
  Let $I$ be finite and $h:I\to\R$. If $h(i_0)=0$ for some $i_0$, then
  $\prod_{i\in I}h(i)=0$. If $h(i)=1$ for every $i$, then
  $\prod_{i\in I}h(i)=1$. These are standard finite-product facts used to turn
  the coordinatewise clipped threshold values into the behavior of the product
  MLP.
\end{lemma}

\begin{lemma}[Scalar division by a product]
  \label{lem:scalar-div-product}
  \lean{div_eq_mul_inv}
  \mathlibok
  For real scalars $s,\alpha,\beta$ with $\alpha\ne0$ and $\beta\ne0$,
  \[
    s\cdot(\alpha^{-1}\beta^{-1})=\frac{s}{\alpha\beta}.
  \]
  This is the elementary field simplification obtained from
  $(\alpha\beta)^{-1}=\alpha^{-1}\beta^{-1}$. In Lean, the corresponding scalar
  goal is closed directly by `field_simp [hα, hβ]` using the nonzero
  hypotheses; there is no subsequent `ring`/`nlinarith` step, because the
  simplifier has already closed the goal in this scalar form.
\end{lemma}

\begin{lemma}[Scaled summand factorization]
  \label{lem:scaled-summand-factor}
  For real scalars $x,y,\alpha,\beta$,
  \[
    (x/\alpha)(y/\beta) = (xy)(\alpha^{-1}\beta^{-1}).
  \]
  This is only the definition of division as multiplication by inverse, followed
  by commutativity and associativity of multiplication. No nonzero denominator
  hypotheses are required for this factorization.
\end{lemma}
\begin{proof}
  Rewrite $x/\alpha$ as $x\alpha^{-1}$ and $y/\beta$ as
  $y\beta^{-1}$. Then reassociate and commute the four scalar factors:
  \[
    (x\alpha^{-1})(y\beta^{-1})
    = x y \alpha^{-1}\beta^{-1}
    = (xy)(\alpha^{-1}\beta^{-1}).
  \]
  In Lean this node should be proved by unfolding division to multiplication by
  inverse and normalizing the commutative monoid product; it should not use
  `field_simp`, because there are no denominator hypotheses and no field
  cancellation is involved.
\end{proof}

\begin{lemma}[Scaling both arguments of the finite-sum inner product]
  \label{lem:inner-scaled}
  \uses{def:inner-prod, lem:scaled-summand-factor, lem:scalar-div-product}
  For any finite index set $I$, vectors $v,w : I \to \R$, and scalars
  $\alpha,\beta \in \R$ with $\alpha\ne 0$ and $\beta\ne 0$,
  \[
    \inner{v/\alpha}{w/\beta}
    =
    \frac{\inner{v}{w}}{\alpha\beta}.
  \]
  Equivalently, after unfolding the inner product as a finite sum,
  \[
    \alpha\beta
    \sum_{i\in I} v_i w_i \alpha^{-1}\beta^{-1}
    =
    \sum_{i\in I} v_i w_i .
  \]
\end{lemma}
\begin{proof}
  \uses{def:inner-prod, lem:scaled-summand-factor, lem:scalar-div-product}
  Expanding the left-hand inner product gives
  \[
    \inner{v/\alpha}{w/\beta}
    =
    \sum_{i\in I} (v_i\alpha^{-1})(w_i\beta^{-1}).
  \]
  For each index $i$, apply Lemma~\ref{lem:scaled-summand-factor}, with
  $x=v_i$ and $y=w_i$, to rewrite the summand as
  \[
    (v_i\alpha^{-1})(w_i\beta^{-1})
    =
    (v_iw_i)(\alpha^{-1}\beta^{-1}).
  \]
  This is a pure reassociation/commutation step and has no nonzero hypotheses;
  in particular the formal proof of this step should use
  Lemma~\ref{lem:scaled-summand-factor}, not a `field_simp` step followed by an
  additional normalizer.
  Since $\alpha^{-1}\beta^{-1}$ is independent of $i$, the finite-sum
  constant-multiple rule gives
  \[
    \sum_{i\in I} (v_iw_i)(\alpha^{-1}\beta^{-1})
    =
    \left(\sum_{i\in I} v_iw_i\right)(\alpha^{-1}\beta^{-1}).
  \]
  Apply Lemma~\ref{lem:scalar-div-product} with
  $s=\sum_{i\in I}v_iw_i$. This rewrites the last display directly as
  $\inner{v}{w}/(\alpha\beta)$ and proves the desired equality. The multiplied
  finite-sum identity displayed in the statement is only the same equality after
  multiplying by the nonzero scalar $\alpha\beta$; it is not an additional step
  needed for the main proof.
\end{proof}

% --- A domain notion that carries logical weight: truly subquadratic time. ---

\begin{definition}[Truly subquadratic time]
  \label{def:truly-subquadratic}
  An algorithm for a problem on inputs of size parameter $n$ comes with a
  concrete worst-case runtime function $T:\N\to\R_{\ge 0}$ (or
  $T:\N\to\N$ embedded in $\R_{\ge0}$). It runs in \emph{truly subquadratic
  time} iff there exist constants $\eps>0$, $C>0$ and $n_0$ such that for every
  $n\ge n_0$,
  \[
    T(n) \le C n^{2-\eps}.
  \]
  For problems with secondary parameters such as $\ell,t,\gamma$, the runtime
  predicate also records the parameter regime and any permitted
  $\operatorname{poly}(\ell)$ factor; the exponent $\eps$ is fixed independently
  of $n$ and the instance. A formal definition that only asserts the existence
  of a positive $\eps$ and does not mention a runtime function $T$ is vacuous and
  is not this node.
\end{definition}

\begin{definition}[Concrete problem families]
  \label{def:problem-family}
  This node introduces the auditable structure
  $\operatorname{ConcreteProblemFamily}$. The declaration checked at this label
  must itself expose the concrete fields below. It may be a Lean `structure', an
  equivalent dependent Sigma package, or a reducible alias/re-export only if
  unfolding the declaration at this exact label exposes the same projections and
  field types. It is not enough for this node to be an abbreviation such as
  \[
    \operatorname{def\_problem\_family} : \mathrm{Type}\;1
      := \operatorname{ConcreteProblemFamily}
  \]
  when the fields of $\operatorname{ConcreteProblemFamily}$ are not visible at
  the audited declaration.

  A \emph{problem family} $\mathcal P$ is concrete data, not a name tag. An
  element $P:\operatorname{ConcreteProblemFamily}$ has:
  \begin{itemize}
    \item a parameter type
      $P.\operatorname{Param}:\mathrm{Type}$;
    \item an input type family
      $P.\operatorname{Inst}:P.\operatorname{Param}\to\mathrm{Type}$;
    \item a kind field
      $P.\operatorname{Kind}\in
      \{\mathrm{search},\mathrm{decision},\mathrm{promise}\}$;
    \item a search-output type family
      $P.\operatorname{Out}:\prod_p
      P.\operatorname{Inst}(p)\to\mathrm{Type}$;
    \item a search-output correctness relation
      $P.\operatorname{ValidOut}:\prod_{p,I}
      P.\operatorname{Out}(p,I)\to\mathrm{Prop}$;
    \item a yes predicate
      $P.\operatorname{Yes}:\prod_p
      P.\operatorname{Inst}(p)\to\mathrm{Prop}$;
    \item a no predicate
      $P.\operatorname{No}:\prod_p
      P.\operatorname{Inst}(p)\to\mathrm{Prop}$.
  \end{itemize}
  For search families, $\operatorname{Out}$ and $\operatorname{ValidOut}$ are
  the nontrivial fields; for decision families, $\operatorname{Yes}$ is the
  nontrivial field; for promise families, both $\operatorname{Yes}$ and
  $\operatorname{No}$ are nontrivial. Unused fields may be filled with a
  canonical unit-valued default, but the projections must still exist with the
  displayed names and types so that later algorithm and reduction packages can
  select their outputs and correctness obligations by $P.\operatorname{Kind}$.

  The parameter tuple is part of the mathematical object: changing $\ell$, $t$
  or $\gamma$ changes the family. For concrete families in this blueprint the
  parameter value $p$ must contain every displayed parameter, such as
  $(n,\ell)$, $(n,\ell,t)$, $(n,\ell,\gamma)$, $(n,\ell,\alpha,\gamma)$, and any
  bichromatic color-class distinction. Subsequent nodes such as $\MSD$, $\OV$
  and $\MaxIP$ must provide unfolding theorems or projection equations
  identifying these fields with their stated input types and relations.

  A formalization that replaces a concrete family by `Unit', a string, an enum,
  a bare tag, an opaque abbreviation, or a declaration that ends in `True' has
  erased the problem and does not satisfy this node. In particular, if the
  generated Lean declares only
  `abbrev def_problem_family : Type 1 := ConcreteProblemFamily', then the
  statement alignment audit must reject it unless this very declaration also
  unfolds to the fields above.
\end{definition}

\begin{definition}[No placeholder problem families]
  \label{def:parameter-sensitive-placeholder}
  \uses{def:problem-family}
  This node introduces, or transparently re-exports, the auditable structure
  $\operatorname{ConcreteProblemFamily}$ used by every reusable problem-family
  wrapper in the blueprint. The mathematical object at this label is a
  structure with the fields below. It may be implemented directly as a
  structure declaration, or as a reducible alias/re-export of an already defined
  structure with exactly these projections; in the alias case, unfolding the
  declaration audited at this label must expose the same field names and field
  types. It is not a marker proposition and it is not an opaque family tag. An
  element $P:\operatorname{ConcreteProblemFamily}$ consists of the following
  fields:
  \[
  \begin{array}{ll}
    P.\operatorname{Param} &: \mathrm{Type},\\
    P.\operatorname{Inst} &: P.\operatorname{Param}\to\mathrm{Type},\\
    P.\operatorname{Kind} &: \{\mathrm{search},\mathrm{decision},\mathrm{promise}\},\\
    P.\operatorname{Out} &: \prod_p P.\operatorname{Inst}(p)\to\mathrm{Type}
      \quad\text{for search families},\\
    P.\operatorname{ValidOut} &: \prod_{p,I}
      P.\operatorname{Out}(p,I)\to\mathrm{Prop}
      \quad\text{for search families},\\
    P.\operatorname{Yes} &: \prod_p P.\operatorname{Inst}(p)\to\mathrm{Prop}
      \quad\text{for decision families},\\
    P.\operatorname{No} &: \prod_p P.\operatorname{Inst}(p)\to\mathrm{Prop}
      \quad\text{for promise families}.
  \end{array}
  \]
  Search families have a nontrivial $\operatorname{Out}$ and
  $\operatorname{ValidOut}$ field; decision families have a nontrivial
  $\operatorname{Yes}$ field; promise families have both nontrivial
  $\operatorname{Yes}$ and $\operatorname{No}$ fields. Unused fields may be
  filled by a canonical unit-valued default, but the kind-specific fields just
  listed are mandatory data of the structure.

  The parameter value $p$ must include every displayed parameter of the problem,
  including $\ell$, $t$, $\gamma$, $\alpha$, and any color-class distinction.
  For each named family below there must be an unfolding theorem identifying
  these fields with the displayed mathematics. For example, unfolding
  the concrete family for $\MSD_{n,\ell,t}$ must give exactly
  \[
    P.\operatorname{Inst}(n,\ell,t)
      =\{(v_1,\ldots,v_n): v_i\in\{0,1\}^{\ell},\ v_i\ne0\},
    \qquad
    P.\operatorname{Yes}(n,\ell,t,I)\Longleftrightarrow
    \exists\, i\ne j,\ \cossim{v_i}{v_j}\ge t.
  \]
  Similarly, unfolding $\gamma$-Min-IP must expose the approximation factor
  $\gamma$ and its two approximation inequalities, and unfolding an additive
  promise problem must expose both its yes and no predicates. A declaration at
  this label whose body is a non-reducible abbreviation, whose target structure
  lacks the displayed projections, whose statement is a bare proposition, which
  ends in `True', which uses `Unit' as the family, or which stores only a
  name/tag/enum plus parameters does not satisfy this node. A transparent alias
  to a previously declared structure is acceptable precisely when field access
  through the audited name reduces to the displayed fields above.
\end{definition}

\begin{definition}[Truly subquadratic algorithm predicate]
  \label{def:tsq-alg-exists}
  \uses{def:problem-family, def:parameter-sensitive-placeholder, def:truly-subquadratic}
  For a concrete problem family $\mathcal P$, this node introduces, or
  transparently re-exports, the auditable structure
  $\operatorname{TSQAlgorithm}(\mathcal P)$. The declaration at this label is
  the structure itself, an equivalent dependent Sigma package with the same
  projections, or a reducible alias to such a package. It is not merely the
  proposition $\operatorname{Nonempty}(\operatorname{TSQAlgorithm}(\mathcal P))$.
  If implemented as an alias, unfolding the audited declaration must expose the
  regime, run, correctness, runtime and exponent fields below. An element
  $\mathcal A:\operatorname{TSQAlgorithm}(\mathcal P)$ has the following fields:
  \begin{itemize}
    \item a parameter-regime predicate
      $\operatorname{Regime}_{\mathcal A} :
      \operatorname{Param}_{\mathcal P}\to\mathrm{Prop}$ specifying the
      asymptotic slice under discussion, such as
      $\ell=\logDim(c,n)$ or $\ell=\msdHardDim(c,n)$;
    \item a single dependent run value, whose type is selected by
      $\mathcal P.\operatorname{Kind}$:
      \[
      \operatorname{AlgOut}_{\mathcal P}(p,I) :=
      \begin{cases}
        \operatorname{Out}_{\mathcal P}(p,I)
          & \text{if }\mathcal P.\operatorname{Kind}=\mathrm{search},\\
        \{\mathrm{true},\mathrm{false}\}
          & \text{if }\mathcal P.\operatorname{Kind}=\mathrm{decision},\\
        \{\mathrm{true},\mathrm{false}\}
          & \text{if }\mathcal P.\operatorname{Kind}=\mathrm{promise}.
      \end{cases}
      \]
      The field is
      \[
        \operatorname{run}_{\mathcal A}(p,I):
        \operatorname{AlgOut}_{\mathcal P}(p,I)
      \]
      for every in-regime $p$ and every instance $I$. There are not separate
      fields called search-run, decision-run and promise-run; requiring all of
      them for every family changes the meaning of the algorithm package.
    \item a single dependent correctness proof, again selected by
      $\mathcal P.\operatorname{Kind}$:
      \[
      \operatorname{AlgCorrect}_{\mathcal P}(p,I,o) :=
      \begin{cases}
        \operatorname{ValidOut}_{\mathcal P}(p,I,o)
          & \text{if }\mathcal P.\operatorname{Kind}=\mathrm{search},\\
        \bigl(o=\mathrm{true}\Longleftrightarrow
          \operatorname{Yes}_{\mathcal P}(p,I)\bigr)
          & \text{if }\mathcal P.\operatorname{Kind}=\mathrm{decision},\\
        \bigl(\operatorname{Yes}_{\mathcal P}(p,I)\Rightarrow
          o=\mathrm{true}\bigr)\wedge
        \bigl(\operatorname{No}_{\mathcal P}(p,I)\Rightarrow
          o=\mathrm{false}\bigr)
          & \text{if }\mathcal P.\operatorname{Kind}=\mathrm{promise}.
      \end{cases}
      \]
      The field is
      \[
        \operatorname{correct}_{\mathcal A}(p,I):
        \operatorname{AlgCorrect}_{\mathcal P}
          (p,I,\operatorname{run}_{\mathcal A}(p,I)).
      \]
      Thus a search algorithm proves only the search output relation; it does
      not also return a Boolean or prove a decision/promise condition. A
      decision or promise algorithm returns only a Boolean; it does not also
      construct a search output. Implementing this node with all three runners
      and all three correctness obligations for every $\mathcal P$ is rejected
      as a different, stronger and semantically misaligned structure.
    \item a worst-case runtime function
      \[
        T_{\mathcal A}:\operatorname{Param}_{\mathcal P}\to\R_{\ge0}
      \]
      bounding the run time on all instances with that parameter;
    \item constants $\eps>0$, $C>0$ and $n_0$ proving that
      $T_{\mathcal A}$ satisfies Definition~\ref{def:truly-subquadratic} on
      $\operatorname{Regime}_{\mathcal A}$, including any allowed
      $\operatorname{poly}(\ell)$ secondary factor and with $\eps$ independent
      of $n$.
  \end{itemize}
  After this structure is defined, the notation
  \[
    \TSQAlg{\mathcal P}
    \quad\text{means}\quad
    \operatorname{Nonempty}(\operatorname{TSQAlgorithm}(\mathcal P)).
  \]
  Thus $\TSQAlg{\MSD_{n,\ell,t}}$ asserts existence of an algorithm for the
  concrete threshold cosine-similarity predicate, not for a family tag named
  ``MSD''. A Lean declaration at this label that merely states
  `Nonempty (TSQAlgorithm P)' is too weak, because the statement being audited
  would not itself expose the regime, run, correctness, runtime function, and
  truly-subquadratic bound data. A reducible type alias to an existing
  `TSQAlgorithm P' structure is acceptable only if these projections are
  available after unfolding the alias. If an implementation uses helper
  definitions such as `AlgOut' and `AlgCorrect', this node must define them,
  import them transparently, or unfold them at this label, and the structure
  must have exactly one kind-indexed run field and exactly one kind-indexed
  correctness field as above.
\end{definition}

\begin{definition}[Truly subquadratic lower-bound predicate]
  \label{def:no-tsq-alg}
  \uses{def:problem-family, def:truly-subquadratic, def:tsq-alg-exists}
  For a problem family $\mathcal P$, the assertion $\NoTSQAlg{\mathcal P}$ is
  definitionally the direct negation of the concrete algorithm package from
  Definition~\ref{def:tsq-alg-exists}:
  \[
    \NoTSQAlg{\mathcal P}
    \quad := \quad
    \neg\,\operatorname{Nonempty}
      \bigl(\operatorname{TSQAlgorithm}(\mathcal P)\bigr).
  \]
  Equivalently, it is the function type
  \[
    \operatorname{TSQAlgorithm}(\mathcal P)\to\mathrm{False}
  \]
  after unpacking `Nonempty'. The definition at this label must mention
  $\operatorname{TSQAlgorithm}(\mathcal P)$ itself and must reduce to its
  negation. It must not negate a separate opaque predicate named `TSQAlg P'
  unless that notation is a transparent abbreviation for the displayed
  `Nonempty (TSQAlgorithm P)'. Otherwise the lower-bound predicate is no longer
  audibly tied to the run, correctness, runtime and exponent fields specified in
  Definition~\ref{def:tsq-alg-exists}.
\end{definition}

\begin{definition}[Fine-grained reduction and equivalence predicates]
  \label{def:fg-reduction-predicates}
  \uses{def:problem-family, def:tsq-alg-exists}
  For concrete problem families $\mathcal P$ and $\mathcal Q$, this node
  introduces, or transparently re-exports, the auditable witness structure
  $\operatorname{FGReduction}(\mathcal P,\mathcal Q)$. The declaration at this
  label is the structure itself, an equivalent dependent Sigma package with the
  same projections, or a reducible alias to such a package. It is not merely the
  proposition
  $\operatorname{Nonempty}(\operatorname{FGReduction}(\mathcal P,\mathcal Q))$.
  If implemented as an alias, unfolding the audited declaration must expose the
  regime, call-list, reconstruction, correctness and runtime-accounting fields
  below. An element $R:\operatorname{FGReduction}(\mathcal P,\mathcal Q)$ has
  fields:
  \begin{itemize}
    \item a source regime
      $\operatorname{SrcRegime}_R:
      \operatorname{Param}_{\mathcal P}\to\mathrm{Prop}$ and a target regime
      $\operatorname{TgtRegime}_R:
      \operatorname{Param}_{\mathcal Q}\to\mathrm{Prop}$, together with an
      eventual-coverage theorem for the displayed source slice. Concretely,
      each reduction states a slice predicate
      $\operatorname{SrcSlice}_R$ and proves
      \[
        \exists n_0,\ \forall p,\ n(p)\ge n_0 \wedge
        \operatorname{SrcSlice}_R(p)
        \Longrightarrow \operatorname{SrcRegime}_R(p).
      \]
      For example, on the logarithmic slice this says that all sufficiently
      large $(n,\ell)$ with $\ell=\logDim(c,n)$ are included. Mere nonemptiness
      of $\operatorname{SrcRegime}_R$ is not enough, and
      $\operatorname{SrcRegime}_R=\emptyset$ is forbidden;
    \item a parameter transformation from each in-regime parameter
      $p_{\mathcal P}$ to a parameter $p_{\mathcal Q}$ of $\mathcal Q$;
    \item a finite oracle-call family, not just a single target instance:
      \[
        \operatorname{Calls}_R(p_{\mathcal P},I):
        \operatorname{Finset}\!
        \left(\sum_{p_{\mathcal Q}:\operatorname{Param}_{\mathcal Q}}
          \operatorname{Inst}_{\mathcal Q}(p_{\mathcal Q})\right).
      \]
      Every call in this finite set must satisfy
      $\operatorname{TgtRegime}_R(p_{\mathcal Q})$. Single-call reductions are
      represented by a singleton finite set; recursive split reductions such as
      Lemma~\ref{lem:minip-bichromatic-reduction} use many calls;
    \item an oracle-answer type for each call, defined by the same kind-indexed
      output convention as Definition~\ref{def:tsq-alg-exists}: a valid target
      search output for search families, and a Boolean answer for decision or
      promise families;
    \item an output reconstruction procedure
      \[
        \operatorname{reconstruct}_R(p_{\mathcal P},I,\operatorname{answers})
        :
        \operatorname{AlgOut}_{\mathcal P}(p_{\mathcal P},I),
      \]
      where `answers' assigns an oracle answer to every element of
      $\operatorname{Calls}_R(p_{\mathcal P},I)$;
    \item a correctness theorem saying that if every oracle answer satisfies
      $\operatorname{AlgCorrect}_{\mathcal Q}$ for its concrete target call,
      then the reconstructed answer satisfies
      $\operatorname{AlgCorrect}_{\mathcal P}$ for the original instance. This
      theorem must use the concrete $\operatorname{ValidOut}$,
      $\operatorname{Yes}$ and $\operatorname{No}$ relations exposed by the two
      problem families;
    \item a runtime accounting package, with named projections and defining
      equations. The structure contains a concrete overhead function
      \[
        \operatorname{overhead}_R(p_{\mathcal P},I)\in\R_{\ge0}
      \]
      for forming the call list and reconstructing the source answer. For every
      target algorithm
      $\mathcal A_Q:\operatorname{TSQAlgorithm}(\mathcal Q)$ it defines the
      oracle answers by the equation
      \[
        \operatorname{oracleAnswer}_{R,\mathcal A_Q}(c)
        =
        \mathcal A_Q.\operatorname{run}(p_c,I_c),
        \qquad
        c=(p_c,I_c)\in\operatorname{Calls}_R(p_{\mathcal P},I),
      \]
      and defines the induced source run by the equation
      \[
        \operatorname{inducedRun}_{R,\mathcal A_Q}(p_{\mathcal P},I)
        =
        \operatorname{reconstruct}_R
          (p_{\mathcal P},I,
           \operatorname{oracleAnswer}_{R,\mathcal A_Q}).
      \]
      Its correctness field is exactly the correctness theorem above applied to
      these oracle answers and to
      $\mathcal A_Q.\operatorname{correct}$ for each concrete call. The induced
      source runtime is also defined, not chosen:
      \[
        \operatorname{inducedRuntime}_{R,\mathcal A_Q}(p_{\mathcal P},I)
        =
        C_R\cdot\operatorname{overhead}_R(p_{\mathcal P},I)
        + C_R\sum_{c=(p_{\mathcal Q},I')\in
          \operatorname{Calls}_R(p_{\mathcal P},I)}
          T_{\mathcal A_Q}(p_{\mathcal Q}),
      \]
      for a fixed constant $C_R>0$ belonging to the reduction witness. The
      runtime-bound theorem states that the runtime of the constructed source
      algorithm is bounded by this very
      $\operatorname{inducedRuntime}_{R,\mathcal A_Q}$, and the
      subquadratic-preservation theorem proves that, whenever
      $\mathcal A_Q$ has exponent $\eps>0$ on the target regime, the induced
      runtime is $O(n^{2-\eps'}\operatorname{poly}(\ell))$ for some
      $\eps'>0$ on the covered source slice. This proof must explicitly account
      for the parameter transformation, dimension blowup, number of calls,
      overhead, and all $\operatorname{poly}(\ell)$ factors.
  \end{itemize}
  The eventual-coverage and call/reconstruction fields are logically essential:
  choosing $\operatorname{SrcRegime}_R=\emptyset$ makes all later obligations
  vacuous, and restricting reductions to one target instance cannot express the
  recursive reductions used in the paper. Every theorem using this structure
  must specify the concrete asymptotic source slice it reduces, and its proof
  object must act on arbitrary sufficiently large instances of that slice.
  The reduction predicate is then
  \[
    \ReducesTo{\mathcal P}{\mathcal Q}
    \quad\Longleftrightarrow\quad
    \operatorname{Nonempty}(\operatorname{FGReduction}(\mathcal P,\mathcal Q)).
  \]
  Every witness canonically induces the logical implication
  $\TSQAlg{\mathcal Q}\Rightarrow \TSQAlg{\mathcal P}$ by building the
  $\operatorname{TSQAlgorithm}(\mathcal P)$ whose run, correctness and runtime
  fields are respectively
  $\operatorname{inducedRun}_{R,\mathcal A_Q}$, the induced correctness theorem,
  and $\operatorname{inducedRuntime}_{R,\mathcal A_Q}$ above. The audited
  declaration must expose this constructor or an equivalent theorem with these
  exact projection equations. There is no free-standing opaque
  `transformedAlgorithm' or `algorithmTransform' field whose behavior can be
  unrelated to the parameter map, call list, target runs, reconstruction,
  correctness proof and runtime theorem. There is also no arbitrary
  `runtimeExpression' field: a nonnegative expression with no equality to the
  target-runtime sum is not runtime accounting. If helper fields with these
  names are present, their equations must reduce to
  $\operatorname{inducedRun}_{R,\mathcal A_Q}$ and
  $\operatorname{inducedRuntime}_{R,\mathcal A_Q}$ as displayed above. The
  implication alone is not the definition of reduction in this blueprint; it
  erases the transformed algorithm, correctness proof, and runtime accounting.
  We write
  $\SubquadEquiv{\mathcal P}{\mathcal Q}$ for the conjunction of the two
  reductions, and $\AtLeastHard{\mathcal Q}{\mathcal P}$ for
  $\ReducesTo{\mathcal P}{\mathcal Q}$. A Lean declaration at this label that
  merely states `Nonempty (FGReduction P Q)' is too weak unless this label is
  also the declaration exposing the fields above. A reducible type alias to an
  existing `FGReduction P Q' package is acceptable only if the projections and
  equations above are available after unfolding the alias. The node is not
  `Unit', not an uninterpreted relation, and not merely the formula
  $\TSQAlg{\mathcal Q}\to\TSQAlg{\mathcal P}$.
\end{definition}

\begin{definition}[Named logarithmic parameter regimes]
  \label{def:log-parameter-regimes}
  \uses{def:problem-family}
  The paper repeatedly uses the same logarithmic dimension and approximation
  regimes. This node must introduce concrete definitions for the six named
  logarithmic-regime functions and predicate. The declaration audited at
  label \ref{def:log-parameter-regimes} is not allowed to be just a value that
  repackages previously existing root constants, and it is not allowed to be a
  structure whose fields are arbitrary functions of the right types.

  A compliant Lean implementation has one of the following two shapes:
  \begin{itemize}
    \item six transparent top-level declarations named exactly
      `logn', `logDim', `polylogDim', `msdHardDim', `mildMsdGamma', and
      `msdHardGammaAllowed', each with the defining equation displayed below,
      and the label declaration \emph{also} exposes those equations by
      transparent simp theorems or projection fields at this same node; or
    \item a single structure/Sigma value named for this label whose data fields
      are the six functions/predicate and whose proof fields state the six
      displayed equations for those very fields.
  \end{itemize}
  In the second shape, the equation fields are mandatory. For example, if
  $R:\operatorname{LogParameterRegimes}$ is the bundled object, the object must
  contain fields such as
  \[
    R.\operatorname{logn\_eq} :
      \forall n,\ R.\operatorname{logn}(n)=\log n
  \]
  and analogous fields for the other five definitions. Merely setting
  $R.\operatorname{logn}:={\_root\_.\logn}$, or storing proofs that separately
  declared constants somewhere else have the desired equations, does not satisfy
  this node; downstream blueprint nodes must be able to unfold or project the
  formulas from the declaration audited here.

  The six required definitions, with domains and defining equations, are:
  \[
    \logn : \N\to\R,\qquad
    \logn(n) = \log n,
  \]
  \[
    \logDim : \R\to\N\to\N,\qquad
    \logDim(c,n) = \left\lceil c\log n\right\rceil,
  \]
  \[
    \polylogDim : \N\to\N\to\N,\qquad
    \polylogDim(k,n) = \left\lceil(\log n)^k\right\rceil,
  \]
  \[
    \msdHardDim : \R\to\N\to\N,\qquad
    \msdHardDim(c,n) =
      \left\lceil(\log n)^{\frac{c\log n}{(\log\log n)^2}}\right\rceil,
  \]
  \[
    \mildMsdGamma : \N\to\R,\qquad
    \mildMsdGamma(n) := 1+\frac{1}{\log\log n}.
  \]
  Finally,
  \[
    \msdHardGammaAllowed : \N\to\R\to\mathrm{Prop}
  \]
  is the predicate
  \[
    \msdHardGammaAllowed(n,\gamma)
    \Longleftrightarrow
    1 \le \gamma \ \wedge\ \gamma \le
    \left(1+\frac{1}{\log\log n}\right)^{\frac{\log n}{(\log\log n)^2}}.
  \]
  Here $\logn$ is real-valued, the dimension functions are natural-valued using
  the fixed ceiling convention, and $\msdHardGammaAllowed$ is a proposition over
  $(n,\gamma)$. Changing ceilings to floors plus a positive lower bound for
  large $n$ would be an equivalent asymptotic convention, but the formal node
  must still provide replacement definitions with the same inputs and outputs.
  The audited declaration must therefore expose all of the following
  field/projection equations, either definitionally or as transparent theorems
  attached to the same declaration:
  \[
  \begin{array}{ll}
    \operatorname{logn\_eq}:&
      \forall n,\ \logn(n)=\log n,\\
    \operatorname{logDim\_eq}:&
      \forall c,n,\ \logDim(c,n)=\lceil c\log n\rceil,\\
    \operatorname{polylogDim\_eq}:&
      \forall k,n,\ \polylogDim(k,n)=\lceil(\log n)^k\rceil,\\
    \operatorname{msdHardDim\_eq}:&
      \forall c,n,\
      \msdHardDim(c,n)=
      \left\lceil(\log n)^{\frac{c\log n}{(\log\log n)^2}}\right\rceil,\\
    \operatorname{mildMsdGamma\_eq}:&
      \forall n,\ \mildMsdGamma(n)=1+\frac{1}{\log\log n},\\
    \operatorname{msdHardGammaAllowed\_iff}:&
      \forall n,\gamma,\ 
      \msdHardGammaAllowed(n,\gamma)\Longleftrightarrow
      1\le\gamma\wedge
      \gamma\le
      \left(1+\frac{1}{\log\log n}\right)^{
        \frac{\log n}{(\log\log n)^2}}.
  \end{array}
  \]
  A declaration at this label is rejected if it has type `Unit', if it ends in
  `True', if it only states a property of $\msdHardGammaAllowed$, if it only
  packages externally existing constants into a value such as
  `def_log_parameter_regimes : LogParameterRegimes', or if its bundled type has
  the six function fields but lacks the six displayed equation fields. The
  formulas must be recoverable from the label declaration itself.
\end{definition}

\chapter{The transformer architecture}

\begin{definition}[Softmax]
  \label{def:softmax}
  This node defines two concrete functions, not just the first one:
  \[
    \operatorname{vectorSoftmax}_n : \R^n\to\R^n,
    \qquad
    \operatorname{matrixSoftmax}_n : \R^{n\times n}\to\R^{n\times n}.
  \]
  For a vector $v \in \R^{n}$,
  \[
    \operatorname{vectorSoftmax}_n(v)
    =
    \frac{(\exp(v[1]), \ldots, \exp(v[n]))}{\sum_{i=1}^{n} \exp(v[i])}
    \in \R^{n}.
  \]
  For a matrix $A \in \R^{n \times n}$ the matrix softmax operator is applied
  row-wise:
  \[
    \operatorname{matrixSoftmax}_n(A)_{i,j}
    =
    \operatorname{vectorSoftmax}_n(A_{i,:})_j
    =
    \frac{\exp(A_{i,j})}{\sum_{k=1}^{n}\exp(A_{i,k})}.
  \]
  We write $\softmax(v)$ and $\softmax(A)$ for these two functions when the
  argument type disambiguates them. A formal declaration at this label must
  expose both the vector function and the row-wise matrix function, together
  with the displayed row equation. A declaration containing only vector softmax
  omits part of Definition~2.1's mathematical input and is not this node.
\end{definition}

\begin{definition}[Attention, Definition 2.1]
  \label{def:attention}
  \uses{def:softmax}
  This node defines the full attention map, not just a container for its
  weights. For positive natural numbers $n,d_{\mathrm{in}},d_{\mathrm{out}},m$,
  an implementation at this label must introduce, or expose as fields with
  projection equations, all of the following objects:
  \[
  \begin{array}{ll}
    Q,K &: \R^{d_{\mathrm{in}}\times m},\\
    V &: \R^{d_{\mathrm{in}}\times d_{\mathrm{out}}},\\
    M_{Q,K} &: \R^{d_{\mathrm{in}}\times d_{\mathrm{in}}},\\
    S_{Q,K} &: \R^{n\times d_{\mathrm{in}}}\to\R^{n\times n},\\
    A_{Q,K,V} &: \R^{n\times d_{\mathrm{in}}}\to
      \R^{n\times d_{\mathrm{out}}}.
  \end{array}
  \]
  The query-key bilinear matrix is definitionally
  \[
    (M_{Q,K})_{a,b}=\sum_{r=1}^m Q_{a,r}K_{b,r}.
  \]
  The score matrix is definitionally the function
  $S_{Q,K}(X) \in \R^{n\times n}$ with entry equation
  \[
    S_{Q,K}(X)_{i,j}
    =
    \sum_{a=1}^{d_{\mathrm{in}}}\sum_{b=1}^{d_{\mathrm{in}}}
      X_{i,a}(M_{Q,K})_{a,b}X_{j,b}.
  \]
  The attention map itself is definitionally the function
  \[
    A_{Q,K,V}(X) = \softmax(S_{Q,K}(X)) X V.
  \]
  Equivalently, its entries satisfy the concrete row-wise softmax formula
  \[
    A_{Q,K,V}(X)_{i,o}
    =
    \sum_{j=1}^{n}
      \frac{\exp(S_{Q,K}(X)_{i,j})}
           {\sum_{k=1}^{n}\exp(S_{Q,K}(X)_{i,k})}
      \left(\sum_{a=1}^{d_{\mathrm{in}}} X_{j,a}V_{a,o}\right).
  \]
  This is exactly the usual matrix formula for pairwise query-key scores, but
  the score and output equations are part of the definition. A Lean declaration
  at this label that merely stores $Q,K,V$, or stores $Q,K,V$ plus an arbitrary
  uninterpreted run function, omits the mathematical content of Definition~2.1.
  If the formalization uses a structure such as `Attention', that structure
  must have transparent projections or simp theorems for `score' and `apply'
  reducing to the two displayed formulas above.
  We write $\mathcal A_{d_{\mathrm{in}}, m, d_{\mathrm{out}}}$ for the set of all
  such attentions.
\end{definition}

\begin{definition}[Multi-layer perceptron, Definition 2.2]
  \label{def:mlp}
  A \emph{multi-layer perceptron} (MLP) is represented by the following concrete
  data:
  \begin{itemize}
    \item natural dimensions $a,b\in\N$;
    \item proofs $0<a$ and $0<b$;
    \item a continuous function $\varphi:\R^{a}\to\R^{b}$.
  \end{itemize}
  The positivity proofs are part of the definition: zero-dimensional source or
  target spaces are not allowed in this blueprint node. This models, following
  the universal approximation theorem, any function approximable by a neural
  network. An MLP is applied to a matrix row-wise: for
  $X \in \R^{n \times a}$,
  \[
    \operatorname{rowApply}_{n,a,b,\varphi}(X)_i=\varphi(X_i),
    \qquad
    \operatorname{rowApply}_{n,a,b,\varphi}(X)\in\R^{n\times b}.
  \]
  A formal `MLP' or `def\_mlp' declaration at this label must therefore store or
  quantify the positive-dimension hypotheses $0<a$ and $0<b$, the continuous
  map, and the row-wise application equation. Allowing arbitrary natural
  dimensions, including $a=0$ or $b=0$, is weaker than this definition.
\end{definition}

\begin{definition}[Flattening matrices for a global output MLP]
  \label{def:matrix-flatten}
  For positive natural numbers $n,d$, fix the standard row-major linear
  equivalence
  \[
    \operatorname{flatten}_{n,d}:\R^{n\times d}\simeq \R^{nd},
    \qquad
    \operatorname{flatten}_{n,d}(Y)_{(i-1)d+r}=Y_{i,r},
  \]
  with inverse $\operatorname{unflatten}_{n,d}$. The equivalence preserves all
  coordinates by the displayed equations and is continuous in both directions.
  This node exists only to make precise how the final MLP of
  Definition~\ref{def:transformer}, whose input is the whole attention-output
  matrix, may be represented as an ordinary vector-input MLP. Any formalization
  that encodes the output MLP as a map $\R^{nd}\to\R$ must include this
  flattening equivalence in the transformer evaluation equation.
\end{definition}

\begin{definition}[Transformer, Definition 2.3]
  \label{def:transformer}
  \uses{def:attention, def:mlp, def:matrix-flatten}
  A \emph{transformer} with one attention unit is the following structured
  object:
  \begin{itemize}
    \item natural dimensions $n,d,d_{\mathrm{in}},m,d_{\mathrm{out}}$ together
      with positivity proofs
      $0<n$, $0<d$, $0<d_{\mathrm{in}}$, $0<m$ and $0<d_{\mathrm{out}}$;
    \item matrices $Q,K\in\R^{d_{\mathrm{in}}\times m}$ and
      $V\in\R^{d_{\mathrm{in}}\times d_{\mathrm{out}}}$ defining the attention
      unit $A_{Q,K,V}$ of Definition~\ref{def:attention};
    \item an input MLP $\varphi_1:\R^d\to\R^{d_{\mathrm{in}}}$ whose stored
      positive dimensions are exactly $d$ and $d_{\mathrm{in}}$, applied
      row-wise to $E\in\R^{n\times d}$ as in Definition~\ref{def:mlp};
    \item an output MLP
      $\varphi_2:\R^{n d_{\mathrm{out}}}\to\R$ whose stored positive dimensions
      are exactly $n d_{\mathrm{out}}$ and $1$.
  \end{itemize}
  The output MLP is global, not row-wise: it receives the entire
  $n\times d_{\mathrm{out}}$ attention-output matrix after applying
  $\operatorname{flatten}_{n,d_{\mathrm{out}}}$. Its run function is not an
  arbitrary field: on an embedding matrix $E \in \R^{n \times d}$ it is
  definitionally
  \[
    \TF(E)
    =
    \varphi_2\!\left(
      \operatorname{flatten}_{n,d_{\mathrm{out}}}
      \left(A_{Q,K,V}(\operatorname{rowApply}_{\varphi_1}(E))\right)
    \right).
  \]
  Thus any formal `Transformer` structure must store the matrices and MLPs (or
  equivalent data from which they can be recovered), the positive-dimension
  hypotheses above, the row-wise application of the input MLP, and the
  flattening equivalence used before the output MLP. Its evaluation equation
  must be the displayed composition. A structure containing only an arbitrary
  function $\R^{n\times d}\to\R$, or an output MLP on $\R^{n d_{\mathrm{out}}}$
  without the displayed flattening/equivalence to $\R^{n\times d_{\mathrm{out}}}$,
  erases architecture that is part of this node.
\end{definition}

\begin{definition}[A transformer solving a decision problem]
  \label{def:tf-solves}
  \uses{def:transformer, def:problem-family}
  Let $\mathcal P$ be a concrete decision problem family, let
  $p:\operatorname{Param}_{\mathcal P}$ be a fixed parameter value, and let
  \[
    \operatorname{embed}_{\mathcal P,p}:
      \operatorname{Inst}_{\mathcal P}(p)\to\R^{n(p)\times d(p)}
  \]
  be the fixed canonical embedding attached to that concrete family and
  parameter. The embedding is only a map from already-valid instances to
  matrices; it does not contain a separate valid-instance predicate and it does
  not redefine the problem semantics. A transformer $\TF$ \emph{solves}
  $\mathcal P$ at the fixed parameter $p$ iff for every instance $I$ of
  $\mathcal P$ at that parameter,
  \[
    \TF(\operatorname{embed}_{\mathcal P,p}(I)) =
    \begin{cases}
      1 & \text{if }\operatorname{Yes}_{\mathcal P}(p,I),\\
      0 & \text{if }\neg\operatorname{Yes}_{\mathcal P}(p,I).
    \end{cases}
  \]
  For the vector problems in this blueprint, $I=(v_1,\ldots,v_n)$ and
  $\operatorname{embed}_{\mathcal P,p}(I)$ is definitionally the matrix whose
  $i$-th row is $v_i$; for document problems it is the matrix of the corresponding
  bag-of-words vectors. This predicate must compare the transformer's output to
  the problem family's own $\operatorname{Yes}$ relation. The tautology
  $\TF(E)=\TF(E)$ is not a solving predicate, and neither is a predicate whose
  truth is controlled by an auxiliary embedding object rather than by
  $\operatorname{Yes}_{\mathcal P}$.
\end{definition}

\begin{definition}[Single-attention-unit realizability predicate]
  \label{def:attention-solves-predicate}
  \uses{def:attention, def:transformer, def:tf-solves, def:problem-family}
  For a concrete decision problem family $\mathcal P$ and a fixed parameter
  value $p$, the assertion $\AttentionSolves{\mathcal P(p)}$ means that there
  exist only the architectural witnesses
  \[
    d_{\mathrm{in}},m,d_{\mathrm{out}},\ Q,K,V,\ \varphi_1,\ \varphi_2
  \]
  with dimensions matching the surrounding theorem, such that the structured
  transformer of Definition~\ref{def:transformer}, built from the full attention
  map of Definition~\ref{def:attention}, solves the concrete yes/no predicate of
  $\mathcal P$ at the already-fixed parameter $p$ in the sense of
  Definition~\ref{def:tf-solves}.

  The parameter $p$, the instance type $\operatorname{Inst}_{\mathcal P}(p)$,
  the yes-predicate $\operatorname{Yes}_{\mathcal P}(p,I)$, and the canonical
  embedding $\operatorname{embed}_{\mathcal P,p}$ are inputs to the predicate,
  not existential witnesses. In particular, the predicate must not existentially
  choose an arbitrary `DecisionEmbedding', an arbitrary valid-instance
  predicate, or a different parameter value. For the named vector decision
  problems the canonical embedding is fixed as the row matrix of the given
  vectors; for $\MSD_{n,\ell,t}$ and $\LSD_{n,\ell,t}$ the fixed parameter is
  exactly $(n,\ell,t)$ and the yes/no relation is exactly the cosine-threshold
  relation from Definitions~\ref{def:msd-dec} and~\ref{def:lsd-dec}.

  A witness therefore consists of matrices and continuous MLPs plus a proof
  \[
    \forall I:\operatorname{Inst}_{\mathcal P}(p),\quad
      \TF(\operatorname{embed}_{\mathcal P,p}(I))=1
      \Longleftrightarrow
      \operatorname{Yes}_{\mathcal P}(p,I),
  \]
  together with the corresponding output-$0$ statement on no instances if the
  formal output is real-valued rather than Boolean. An arbitrary function, a
  constant function, a proof using only an uninterpreted problem tag, or a proof
  made vacuous by choosing an empty auxiliary valid-instance predicate does not
  witness $\AttentionSolves{\mathcal P(p)}$.
\end{definition}

\chapter{Document similarity embeddings and problems}

\begin{definition}[Bag-of-words embedding]
  \label{def:bow}
  A document is modeled as a finite sequence of tokens over a token type
  $\Sigma$. For a fixed keyword list
  $K:\{1,\ldots,\ell\}\to\Sigma$, define the occurrence predicate
  \[
    \operatorname{Occurs}(D,K_i)
    \Longleftrightarrow
    \exists r,\ D[r]=K_i .
  \]
  The \emph{bag-of-words embedding} is the concrete function
  \[
    \BOW_K : \operatorname{Document}(\Sigma)\to\{0,1\}^{\ell}
  \]
  defined coordinatewise by
  \[
    \BOW_K(D)[i] =
    \begin{cases}
      1 & \text{if }\operatorname{Occurs}(D,K_i),\\
      0 & \text{otherwise.}
    \end{cases}
  \]
  Equivalently, for every document $D$ and every coordinate $i$,
  \[
    \BOW_K(D)[i]=1
    \Longleftrightarrow
    \operatorname{Occurs}(D,K_i),
    \qquad
    \BOW_K(D)[i]=0
    \Longleftrightarrow
    \neg\operatorname{Occurs}(D,K_i).
  \]
  The formal declaration at this label must therefore provide the function
  $\BOW_K$ and the coordinate iff theorem above. A structure that stores only a
  vocabulary, token data, or an occurrence relation without defining the output
  vector and its $0/1$ coordinate behavior is not the bag-of-words embedding.
  For example, a document containing only "There are ten apples on the apple
  tree", with keywords "apple", "tree", "computer", "ten", has embedding
  $(1,1,0,1)$.
\end{definition}

\begin{definition}[Cosine similarity]
  \label{def:cosine}
  \uses{def:inner-prod, def:l2norm}
  For nonzero document embeddings $v, w \in \R^{d}$, their \emph{cosine
  similarity} is
  \[
    \cossim{v}{w} = \frac{\inner{v}{w}}{\norm{v}\norm{w}}.
  \]
  This value lies in $[0,1]$ when $v,w \in \{0,1\}^d$; the value $1$ means
  complete similarity and $0$ means no similarity. (Used instead of the raw inner
  product so that two vectors with similar directions but large magnitude are
  still considered close.)
\end{definition}

\begin{lemma}[Cosine as inner product after Euclidean normalization]
  \label{lem:cosine-normalized-dot}
  \uses{def:cosine, def:inner-prod, def:l2norm, lem:inner-scaled}
  For nonzero vectors $v,w \in \R^\ell$, if
  $\widehat v = v/\norm{v}$ and $\widehat w = w/\norm{w}$, then
  $\inner{\widehat v}{\widehat w} = \cossim{v}{w}$.
\end{lemma}
\begin{proof}
  \uses{def:cosine, def:inner-prod, def:l2norm, lem:inner-scaled}
  Apply Lemma~\ref{lem:inner-scaled} with
  $\alpha=\norm{v}$ and $\beta=\norm{w}$. The hypotheses that $v$ and $w$ are
  nonzero give $\norm{v}\ne0$ and $\norm{w}\ne0$, so
  \[
    \inner{v/\norm{v}}{w/\norm{w}}
    =
    \frac{\inner{v}{w}}{\norm{v}\norm{w}}.
  \]
  The right-hand side is exactly $\cossim{v}{w}$ by
  Definition~\ref{def:cosine}.
\end{proof}

% --- Most/Least Similar Document problems and their variants. ---------------

\begin{definition}[Most Similar Document, $\MSD_{n,\ell}$, Definition 2.10]
  \label{def:msd}
  \uses{def:cosine, def:parameter-sensitive-placeholder}
  For parameters $n,\ell \in \N$, an instance of $\MSD_{n,\ell}$ is an
  $n$-tuple $(v_1,\ldots,v_n)$ of nonzero vectors in $\{0,1\}^{\ell}$. A valid
  output is a pair $(i,j)$ with $1 \le i,j \le n$, $i \ne j$, such that
  \[
    \cossim{v_i}{v_j}
    =
    \max_{\substack{1\le p,q\le n\\p\ne q}}\cossim{v_p}{v_q}.
  \]
  Formally, this node defines the input predicate ``all rows are nonzero binary
  vectors'', the output type of distinct index pairs, and the
  $\operatorname{ValidOut}$ relation above. Note $\MSD$ differs from $\MaxIP$
  because of the normalization by $\norm{v_i}\norm{v_j}$. A declaration that is
  only a parameter tag named $\MSD$ omits the search relation and is not aligned
  with this definition.
\end{definition}

\begin{definition}[$\gamma$-MSD, Definition 2.11]
  \label{def:gamma-msd}
  \uses{def:cosine, def:parameter-sensitive-placeholder}
  For parameters $n,\ell\in\N$ and approximation factor $\gamma\ge 1$, an
  instance of $\gamma\text{-}\MSD_{n,\ell}$ is an $n$-tuple of nonzero vectors
  $v_1,\ldots,v_n\in\{0,1\}^{\ell}$. A valid output is a pair
  $i^{*}, j^{*}$ with $i^{*}\ne j^{*}$ such that
  \[
    \frac{1}{\gamma}\cdot
    \max_{\substack{1 \le i,j \le n\\i\ne j}} \cossim{v_i}{v_j}
    \le \cossim{v_{i^{*}}}{v_{j^{*}}}
    \le
    \max_{\substack{1 \le i,j \le n\\i\ne j}} \cossim{v_i}{v_j}.
  \]
  The premise $\gamma\ge1$ and the two displayed inequalities are part of the
  formal output relation. A formalization must encode the approximation factor,
  the maximum over distinct pairs, and the returned pair; it cannot replace the
  problem by an uninterpreted family tag.
\end{definition}

\begin{definition}[MSD decision version, $\MSD_{n,\ell,t}$, Definition 2.12]
  \label{def:msd-dec}
  \uses{def:cosine, def:parameter-sensitive-placeholder}
  For parameters $n,\ell\in\N$ and threshold $t\in[0,1]$, an instance of
  $\MSD_{n,\ell,t}$ is an $n$-tuple of nonzero vectors
  $v_1,\ldots,v_n\in\{0,1\}^{\ell}$. It is a yes instance iff there exist
  $1\le i,j\le n$ with $i\ne j$ and $\cossim{v_i}{v_j} \ge t$. Thus the
  yes-predicate is exactly
  \[
    \exists\, i\ne j,\quad \cossim{v_i}{v_j}\ge t.
  \]
  The threshold $t$ and the nonzero-binary input promise are part of the
  definition. This node is a concrete Boolean relation, not a placeholder tag.
\end{definition}

\begin{definition}[Least Similar Document, $\LSD_{n,\ell}$, Definition A.14]
  \label{def:lsd}
  \uses{def:cosine, def:parameter-sensitive-placeholder}
  For parameters $n,\ell\in\N$, an instance of $\LSD_{n,\ell}$ is an $n$-tuple
  of nonzero vectors $v_1,\ldots,v_n\in\{0,1\}^{\ell}$. A valid output is a pair
  $(i,j)$ with $i\ne j$ minimizing $\cossim{v_i}{v_j}$ over all distinct pairs:
  \[
    \cossim{v_i}{v_j}
    =
    \min_{\substack{1\le p,q\le n\\p\ne q}}\cossim{v_p}{v_q}.
  \]
  Formally this is a search relation over nonzero binary vectors and distinct
  output indices, not a tag.
\end{definition}

\begin{definition}[$\gamma$-LSD, Definition A.15]
  \label{def:gamma-lsd}
  \uses{def:cosine, def:parameter-sensitive-placeholder}
  For parameters $n,\ell\in\N$ and approximation factor $\gamma\ge 1$, an
  instance of $\gamma\text{-}\LSD_{n,\ell}$ is an $n$-tuple of nonzero vectors
  $v_1,\ldots,v_n\in\{0,1\}^{\ell}$. A valid output is a pair
  $i^{*} \ne j^{*}$ such that
  \[
    \min_{\substack{1 \le i,j \le n\\i\ne j}} \cossim{v_i}{v_j}
    \le \cossim{v_{i^{*}}}{v_{j^{*}}}
    \le \gamma \cdot
    \min_{\substack{1 \le i,j \le n\\i\ne j}} \cossim{v_i}{v_j}.
  \]
  The formal output relation must include the premise $\gamma\ge1$, the minimum
  over distinct pairs, and the two approximation inequalities. It cannot be
  represented by a bare family tag.
\end{definition}

\begin{definition}[LSD decision version, $\LSD_{n,\ell,t}$, Definition A.16]
  \label{def:lsd-dec}
  \uses{def:cosine, def:parameter-sensitive-placeholder}
  For parameters $n,\ell\in\N$ and threshold $t\in[0,1]$, an instance of
  $\LSD_{n,\ell,t}$ is an $n$-tuple of nonzero vectors
  $v_1,\ldots,v_n\in\{0,1\}^{\ell}$. It is a yes instance iff there exist
  $1\le i,j\le n$ with $i\ne j$ and $\cossim{v_i}{v_j} \le t$. Thus the
  yes-predicate is exactly
  \[
    \exists\, i\ne j,\quad \cossim{v_i}{v_j}\le t,
  \]
  with the nonzero-binary input promise and threshold $t$ included in the
  relation.
\end{definition}

\begin{definition}[Bichromatic $\gamma$-LSD]
  \label{def:bichromatic-gamma-lsd}
  \uses{def:cosine, def:parameter-sensitive-placeholder}
  For parameters $n,\ell\in\N$ and $\gamma\ge1$, an instance of bichromatic
  $\gamma\text{-}\LSD_{n,\ell}$ is a pair of $n$-tuples
  $A=(a_1,\ldots,a_n)$ and $B=(b_1,\ldots,b_n)$ of nonzero vectors in
  $\{0,1\}^{\ell}$. A valid output is a cross-color pair $(i,j)$ with
  $1\le i,j\le n$ such that, writing
  \[
    c_I(p,q)=\cossim{a_p}{b_q},
    \qquad
    m_I=\min_{1\le p,q\le n} c_I(p,q),
  \]
  one has
  \[
    m_I \le c_I(i,j) \le \gamma m_I .
  \]
  The output indices range over different color classes, so no $i\ne j$
  condition is present. The nonzero-binary promises on both lists, the
  approximation factor $\gamma$, the cross-color minimum $m_I$, and the two
  approximation inequalities are the concrete search relation. A formalization
  that treats this as the monochromatic $\gamma$-LSD family with a tag, or that
  omits the $A\times B$ indexing, is not this problem.
\end{definition}

\chapter{Fine-grained complexity: hypotheses and intermediate problems}

% --- The hardness hypotheses, stated as assumptions/conjectures. ------------

\begin{definition}[$k$-SAT]
  \label{def:ksat}
  \uses{def:parameter-sensitive-placeholder}
  For a parameter $k\in\N$, $k\SAT$ is the satisfiability problem for Boolean
  formulas in conjunctive normal form in which every clause has at most
  $k$ literals; $n$ denotes the number of variables. An instance is such a CNF
  formula, and it is a yes instance iff there exists a Boolean assignment of its
  $n$ variables satisfying every clause. Recalled to state SETH. This is a
  concrete decision problem, not merely the symbol $k\SAT$.
\end{definition}

\begin{assumption}[Strong Exponential Time Hypothesis (SETH), Definition 2.4 / A.1]
  \label{assumption:seth}
  \uses{def:ksat, def:truly-subquadratic}
  A $k\SAT$ algorithm is concrete data: a run function on $k$-CNF formulas, a
  proof that the returned Boolean is true exactly on satisfiable formulas, and a
  worst-case runtime function $T_{\mathcal A}(n)$ over formulas with $n$
  variables. SETH is the proposition
  \[
    \forall\,\eps>0,\ \exists\,k\ge1,\ \neg\exists\,\mathcal A,\ C>0,\ n_0,\
      \forall n\ge n_0,\quad
        T_{\mathcal A}(n) \le C\cdot 2^{(1-\eps)n}.
  \]
  Equivalently, for every exponent saving $\eps>0$ there is a clause width
  $k$ for which no correct algorithm decides concrete $k\SAT$ in time
  $O(2^{(1-\eps)n})$. The quantified lower-bound content, the algorithm
  correctness relation for satisfiability, the runtime function, and the
  exponential bound are all part of this assumption. A statement that merely
  quantifies $\eps,k,n$ and concludes `True' for large $n$ is not SETH.
\end{assumption}

\begin{definition}[Orthogonal Vectors, $\OV_{n,\ell}$, Definition 2.5 / A.2]
  \label{def:ov}
  \uses{def:inner-prod, def:parameter-sensitive-placeholder}
  For parameters $n,\ell\in\N$, an instance of $\OV_{n,\ell}$ is an $n$-tuple
  $v_1, \ldots, v_n \in \{0,1\}^{\ell}$. It is a yes instance iff there exists a
  pair $1\le i,j\le n$ with $i\ne j$ and $\inner{v_i}{v_j} = 0$. Thus the
  formal yes-predicate is
  \[
    \exists\, i\ne j,\quad \sum_{r=1}^{\ell} v_i[r]v_j[r]=0.
  \]
  The input type, binary promise, distinctness condition and orthogonality
  equation must all be encoded.
\end{definition}

\begin{definition}[Bichromatic Orthogonal Vectors]
  \label{def:bichromatic-ov}
  \uses{def:inner-prod, def:parameter-sensitive-placeholder}
  For parameters $n,\ell\in\N$, an instance of bichromatic $\OV_{n,\ell}$ is a
  pair of $n$-tuples $A = \{a_1, \ldots, a_n\}$ and
  $B = \{b_1, \ldots, b_n\}$ of vectors in $\{0,1\}^{\ell}$. It is a yes
  instance iff there exist $i,j$ with $\inner{a_i}{b_j} = 0$. Unlike the
  monochromatic version, $i$ and $j$ range over different color classes and need
  not be distinct. This concrete two-input yes-predicate is the content of the
  node.
\end{definition}

\begin{conjecture}[Orthogonal Vectors Conjecture (OVC), Conjecture 2.6 / A.3]
  \label{conjecture:ovc}
  \uses{def:ov, def:no-tsq-alg, def:log-parameter-regimes}
  For any $\eps > 0$ there exists a constant $c > 0$ such that
  $\NoTSQAlg{\OV_{n,\logDim(c,n)}}$; equivalently, the dimension is
  $c\log n$ up to the fixed rounding convention.
\end{conjecture}

\begin{assumption}[SETH implies OVC, Williams (2005)]
  \label{thm:seth-implies-ovc}
  \uses{assumption:seth, conjecture:ovc}
  This is a cited external dependency from Williams (2005), not a proof
  reproduced in the paper. Its mathematical content is the implication
  \[
    \text{SETH}\Longrightarrow \text{OVC}.
  \]
  Consequently all hardness results below can be stated under SETH \emph{or}
  OVC. A formalization of this blueprint may register this node as an axiom or
  imported cited theorem. It should not generate a local proof obligation unless
  it also formalizes the standard split-and-list reduction from $k$-SAT to
  Orthogonal Vectors and the corresponding runtime calculation.
\end{assumption}

% --- Minimum Inner Product problems. ---------------------------------------

\begin{definition}[Minimum Inner Product, $\MinIP_{n,\ell}$, Definition 2.7 / A.5]
  \label{def:minip}
  \uses{def:inner-prod, def:parameter-sensitive-placeholder}
  For parameters $n,\ell\in\N$, an instance of $\MinIP_{n,\ell}$ is an $n$-tuple
  $v_1, \ldots, v_n \in \{0,1\}^{\ell}$. A valid output is a distinct pair
  $(i,j)$ minimizing $\inner{v_i}{v_j}$ over all distinct pairs:
  \[
    \inner{v_i}{v_j} =
    \min_{\substack{1\le p,q\le n\\p\ne q}}\inner{v_p}{v_q}.
  \]
  This is a search relation over binary vectors and distinct indices.
\end{definition}

\begin{definition}[$\gamma$-Min-IP, Definition 2.8 / A.6]
  \label{def:gamma-minip}
  \uses{def:inner-prod, def:parameter-sensitive-placeholder}
  For parameters $n,\ell\in\N$ and $\gamma\ge 1$, an instance of
  $\gamma\text{-}\MinIP_{n,\ell}$ is an $n$-tuple
  $v_1, \ldots, v_n \in \{0,1\}^{\ell}$. A valid output is a distinct pair
  $(i,j)$ satisfying the following relation. Define the score and optimum
  \[
    s_I(p,q)=\inner{v_p}{v_q},
    \qquad
    m_I=\min_{\substack{p,q\\p\ne q}}s_I(p,q).
  \]
  Then
  \[
    \operatorname{GammaMinIPValidOut}(n,\ell,\gamma,I,(i,j))
    \Longleftrightarrow
    i\ne j\ \wedge\ \gamma\ge1\ \wedge\
    m_I\le s_I(i,j)\ \wedge\ s_I(i,j)\le\gamma m_I .
  \]
  The premise $\gamma\ge1$ and this multiplicative approximation relation are
  part of the definition. The output score is allowed to be larger than the
  minimum by a factor of at most $\gamma$; it is not required to dominate all
  scores. A declaration that states
  $\forall p,q,\ s_I(p,q)\le s_I(i,j)$, or
  $s_I(i,j)\le\gamma s_I(p,q)$ for every pair $(p,q)$, is a Max-IP-style
  relation and does not satisfy this node. The audited declaration must expose
  the named score $s_I$, optimum $m_I$, and the displayed iff for
  $\operatorname{GammaMinIPValidOut}$.
\end{definition}

\begin{definition}[Min-IP decision version, $\MinIP_{n,\ell,t}$, Definition 2.9 / A.7]
  \label{def:minip-dec}
  \uses{def:inner-prod, def:parameter-sensitive-placeholder}
  For parameters $n,\ell\in\N$ and $0\le t\le \ell$, an instance of
  $\MinIP_{n,\ell,t}$ is an $n$-tuple $v_1,\ldots,v_n\in\{0,1\}^{\ell}$. It is a
  yes instance iff there exists a distinct pair $(i,j)$ with
  $\inner{v_i}{v_j}\le t$. The threshold $t$ and this existential inequality are
  part of the yes-predicate.
\end{definition}

\begin{definition}[Bichromatic Min-IP]
  \label{def:bichromatic-minip}
  \uses{def:inner-prod, def:parameter-sensitive-placeholder}
  This node defines the whole bichromatic Min-IP family bundle at this label.
  The declaration audited at \ref{def:bichromatic-minip} must expose all four
  named components below, either as fields of a structure
  $\operatorname{BichromaticMinIPBundle}$ or as an equivalent dependent Sigma
  package. A declaration for only the exact search family is not this node.

  First, the shared instance type is
  \[
    \operatorname{BMinIPInst}(n,\ell)
    =
    \{(A,B): A=(a_1,\ldots,a_n),\ B=(b_1,\ldots,b_n),\
      a_i,b_j\in\{0,1\}^{\ell}\}.
  \]
  For $I=(A,B)$ define
  \[
    s_I(i,j)=\inner{a_i}{b_j},\qquad
    m_I=\min_{1\le p,q\le n}s_I(p,q).
  \]
  The bundle then contains three separate concrete problem-family declarations:
  \[
  \begin{array}{ll}
    \operatorname{BMinIPExact}:&
      \operatorname{Param}=\N\times\N,\quad
      \operatorname{Inst}(n,\ell)=\operatorname{BMinIPInst}(n,\ell),\\
    &
      \operatorname{Out}(n,\ell,I)=
        \{(i,j):1\le i,j\le n\},\\
    &
      \operatorname{ValidOut}(n,\ell,I,(i,j))
        \Longleftrightarrow s_I(i,j)=m_I;\\[2mm]
    \operatorname{BMinIPApprox}:&
      \operatorname{Param}=\N\times\N\times\R,\quad
      \operatorname{Inst}(n,\ell,\gamma)=\operatorname{BMinIPInst}(n,\ell),\\
    &
      \operatorname{Out}(n,\ell,\gamma,I)=
        \{(i,j):1\le i,j\le n\},\\
    &
      \operatorname{ValidOut}_{\gamma}(n,\ell,\gamma,I,(i,j))
        \Longleftrightarrow
        \gamma\ge1\ \wedge\ m_I\le s_I(i,j)\le \gamma m_I;\\[2mm]
    \operatorname{BMinIPThreshold}:&
      \operatorname{Param}=\N\times\N\times\R,\quad
      \operatorname{Inst}(n,\ell,t)=\operatorname{BMinIPInst}(n,\ell),\\
    &
      \operatorname{Yes}_{t}(n,\ell,t,I)
        \Longleftrightarrow
        \exists\, i,j,\ 1\le i,j\le n\ \wedge\ s_I(i,j)\le t.
  \end{array}
  \]
  Here $\operatorname{BMinIPExact}$ is a search family with parameter
  $(n,\ell)$, $\operatorname{BMinIPApprox}$ is a different search family with
  parameter $(n,\ell,\gamma)$, and $\operatorname{BMinIPThreshold}$ is a
  decision family with parameter $(n,\ell,t)$. The approximation factor
  $\gamma$ and threshold $t$ are not optional metadata and cannot share a single
  overloaded parameter. The exact and approximate output relations must not be
  merged by a disjunction. In particular,
  $\operatorname{BMinIPApprox}$ must be definitionally tied to the minimum
  $m_I$: it is rejected if it requires
  $\forall p,q,\ s_I(p,q)\le s_I(i,j)$, if it requires
  $s_I(i,j)\le\gamma s_I(p,q)$ for every pair, or if it otherwise makes the
  returned pair maximal rather than within a factor $\gamma$ of the minimum.
  The statement at this label must include the projections or unfolding
  theorems for
  $\operatorname{BMinIPInst}$, $\operatorname{BMinIPExact}$,
  $\operatorname{BMinIPApprox}$, and $\operatorname{BMinIPThreshold}$; otherwise
  later uses of the approximate and threshold bichromatic problems have no
  concrete semantics.
\end{definition}

% --- Maximum Inner Product problems. ---------------------------------------

\begin{definition}[Maximum Inner Product, $\MaxIP_{n,\ell}$, Definition A.9]
  \label{def:maxip}
  \uses{def:inner-prod, def:parameter-sensitive-placeholder}
  For parameters $n,\ell\in\N$, an instance of $\MaxIP_{n,\ell}$ is an $n$-tuple
  $v_1, \ldots, v_n \in \{0,1\}^{\ell}$. A valid output is a distinct pair
  $(i,j)$ maximizing $\inner{v_i}{v_j}$ over all distinct pairs:
  \[
    \inner{v_i}{v_j} =
    \max_{\substack{1\le p,q\le n\\p\ne q}}\inner{v_p}{v_q}.
  \]
  This is a concrete search relation over binary vectors and distinct indices.
\end{definition}

\begin{definition}[$\gamma$-Max-IP, Definition A.10]
  \label{def:gamma-maxip}
  \uses{def:inner-prod, def:parameter-sensitive-placeholder}
  For parameters $n,\ell\in\N$ and $\gamma\ge 1$, an instance of
  $\gamma\text{-}\MaxIP_{n,\ell}$ is an $n$-tuple
  $v_1, \ldots, v_n \in \{0,1\}^{\ell}$. A valid output is a distinct pair whose
  inner product is a $\gamma$-approximation of the maximal inner product.
  Define
  \[
    s_I(p,q)=\inner{v_p}{v_q},
    \qquad
    M_I=\max_{\substack{p,q\\p\ne q}}s_I(p,q).
  \]
  Then
  \[
    \operatorname{GammaMaxIPValidOut}(n,\ell,\gamma,I,(i,j))
    \Longleftrightarrow
    i\ne j\ \wedge\ \gamma\ge1\ \wedge\
    \gamma^{-1}M_I\le s_I(i,j)\ \wedge\ s_I(i,j)\le M_I .
  \]
  The factor $\gamma$ and the premise $\gamma\ge1$ are part of the output
  relation. For $\gamma>1$ this is intentionally weaker than exact
  maximization: any pair whose score lies in the displayed interval is valid.
  A declaration that requires $\forall p,q,\ s_I(p,q)\le s_I(i,j)$ is the exact
  $\MaxIP$ relation, not this approximate relation, even though exact
  maximizers are valid examples. The audited declaration must expose the named
  score $s_I$, optimum $M_I$, and the displayed iff for
  $\operatorname{GammaMaxIPValidOut}$.
\end{definition}

\begin{definition}[Max-IP decision version, $\MaxIP_{n,\ell,t}$, Definition A.11]
  \label{def:maxip-dec}
  \uses{def:inner-prod, def:parameter-sensitive-placeholder}
  For parameters $n,\ell\in\N$ and $0\le t\le \ell$, an instance of
  $\MaxIP_{n,\ell,t}$ is an $n$-tuple $v_1,\ldots,v_n\in\{0,1\}^{\ell}$. It is a
  yes instance iff there exists a distinct pair $(i,j)$ with
  $\inner{v_i}{v_j}\ge t$. The threshold and this existential inequality are the
  concrete yes-predicate.
\end{definition}

\begin{definition}[Bichromatic Max-IP]
  \label{def:bichromatic-maxip}
  \uses{def:inner-prod, def:parameter-sensitive-placeholder}
  This node defines the whole bichromatic Max-IP family bundle at this label.
  The declaration audited at \ref{def:bichromatic-maxip} must expose all four
  named components below, either as fields of a structure
  $\operatorname{BichromaticMaxIPBundle}$ or as an equivalent dependent Sigma
  package. A declaration for only the exact search family is not this node.

  First, the shared instance type is
  \[
    \operatorname{BMaxIPInst}(n,\ell)
    =
    \{(A,B): A=(a_1,\ldots,a_n),\ B=(b_1,\ldots,b_n),\
      a_i,b_j\in\{0,1\}^{\ell}\}.
  \]
  For $I=(A,B)$ define
  \[
    s_I(i,j)=\inner{a_i}{b_j},\qquad
    M_I=\max_{1\le p,q\le n}s_I(p,q).
  \]
  The bundle then contains three separate concrete problem-family declarations:
  \[
  \begin{array}{ll}
    \operatorname{BMaxIPExact}:&
      \operatorname{Param}=\N\times\N,\quad
      \operatorname{Inst}(n,\ell)=\operatorname{BMaxIPInst}(n,\ell),\\
    &
      \operatorname{Out}(n,\ell,I)=
        \{(i,j):1\le i,j\le n\},\\
    &
      \operatorname{ValidOut}(n,\ell,I,(i,j))
        \Longleftrightarrow s_I(i,j)=M_I;\\[2mm]
    \operatorname{BMaxIPApprox}:&
      \operatorname{Param}=\N\times\N\times\R,\quad
      \operatorname{Inst}(n,\ell,\gamma)=\operatorname{BMaxIPInst}(n,\ell),\\
    &
      \operatorname{Out}(n,\ell,\gamma,I)=
        \{(i,j):1\le i,j\le n\},\\
    &
      \operatorname{ValidOut}_{\gamma}(n,\ell,\gamma,I,(i,j))
        \Longleftrightarrow
        \gamma\ge1\ \wedge\ \gamma^{-1}M_I\le s_I(i,j)\le M_I;\\[2mm]
    \operatorname{BMaxIPThreshold}:&
      \operatorname{Param}=\N\times\N\times\R,\quad
      \operatorname{Inst}(n,\ell,t)=\operatorname{BMaxIPInst}(n,\ell),\\
    &
      \operatorname{Yes}_{t}(n,\ell,t,I)
        \Longleftrightarrow
        \exists\, i,j,\ 1\le i,j\le n\ \wedge\ s_I(i,j)\ge t.
  \end{array}
  \]
  Here $\operatorname{BMaxIPExact}$ is a search family with parameter
  $(n,\ell)$, $\operatorname{BMaxIPApprox}$ is a different search family with
  parameter $(n,\ell,\gamma)$, and $\operatorname{BMaxIPThreshold}$ is a
  decision family with parameter $(n,\ell,t)$. The approximation factor
  $\gamma$ and threshold $t$ are not optional metadata and cannot share a single
  overloaded parameter. The exact and approximate output relations must not be
  merged by a disjunction. In particular,
  $\operatorname{BMaxIPApprox}$ must be definitionally tied to the interval
  $\gamma^{-1}M_I\le s_I(i,j)\le M_I$: it is rejected if it requires the exact
  maximizer condition $\forall p,q,\ s_I(p,q)\le s_I(i,j)$ as its validity
  relation. Exact maximizers satisfy the approximate relation, but the
  approximate component's $\operatorname{ValidOut}_{\gamma}$ must be exactly the
  displayed approximation relation, not the stronger exact one. The statement at
  this label must include the projections or unfolding theorems for
  $\operatorname{BMaxIPInst}$, $\operatorname{BMaxIPExact}$,
  $\operatorname{BMaxIPApprox}$, and $\operatorname{BMaxIPThreshold}$; otherwise
  later uses of the approximate and threshold bichromatic problems have no
  concrete semantics.
\end{definition}

\chapter{Fine-grained reductions between intermediate problems}

\begin{lemma}[Append-coordinate construction for bichromatic OV]
  \label{lem:ov-bichromatic-append-construction}
  \uses{def:bichromatic-ov, def:ov, def:inner-prod}
  Given a bichromatic $\OV_{n,\ell}$ instance
  $A=(a_1,\ldots,a_n)$ and $B=(b_1,\ldots,b_n)$, define
  \[
    a_i'=(a_i,1,0)\in\{0,1\}^{\ell+2},
    \qquad
    b_j'=(b_j,0,1)\in\{0,1\}^{\ell+2},
  \]
  and let $W=(a_1',\ldots,a_n',b_1',\ldots,b_n')$ be the corresponding
  monochromatic $\OV_{2n,\ell+2}$ instance. The construction has the following
  concrete behavior:
  \[
    \inner{a_i'}{a_{i'}'}=\inner{a_i}{a_{i'}}+1\ge1,\qquad
    \inner{b_j'}{b_{j'}'}=\inner{b_j}{b_{j'}}+1\ge1,
  \]
  and
  \[
    \inner{a_i'}{b_j'}=\inner{a_i}{b_j}.
  \]
  Consequently,
  \[
    \operatorname{Yes}_{\mathrm{BOV}}(A,B)
    \Longleftrightarrow
    \operatorname{Yes}_{\OV}(W).
  \]
  The formal witness at this label must include the actual maps
  $a_i\mapsto a_i'$ and $b_j\mapsto b_j'$, the three displayed inner-product
  equalities/inequalities, and the final yes-predicate equivalence. Merely
  asserting that some transformed instance exists, or that the construction is
  a reduction, omits the behavior needed by Lemma~\ref{lem:ov-bichromatic-equiv}.
\end{lemma}
\begin{proof}
  \uses{def:bichromatic-ov, def:ov, def:inner-prod}
  Expanding the inner product in the last two coordinates gives
  \[
    \inner{(a_i,1,0)}{(a_{i'},1,0)}
      =\inner{a_i}{a_{i'}}+1,
    \qquad
    \inner{(b_j,0,1)}{(b_{j'},0,1)}
      =\inner{b_j}{b_{j'}}+1,
  \]
  so two vectors from the same color class in the transformed monochromatic
  instance can never be orthogonal. Similarly,
  \[
    \inner{(a_i,1,0)}{(b_j,0,1)}
      =\inner{a_i}{b_j}+1\cdot0+0\cdot1
      =\inner{a_i}{b_j}.
  \]
  Therefore the only orthogonal pairs in $W$ are exactly transformed
  cross-color pairs, and the bichromatic and monochromatic yes predicates are
  equivalent.
\end{proof}

\begin{lemma}[OV is subquadratic equivalent to bichromatic OV, Lemma A.4]
  \label{lem:ov-bichromatic-equiv}
  \uses{def:ov, def:bichromatic-ov, lem:ov-bichromatic-append-construction, def:fg-reduction-predicates, def:tsq-alg-exists}
  This node asserts a concrete pair of fine-grained reductions, not merely an
  abstract logical equivalence:
  \[
    \SubquadEquiv{\OV_{n,\ell}}{\text{bichromatic }\OV_{n,\ell}}.
  \]
  The declaration audited at this label must be the concrete witness package
  $\operatorname{OVBichromaticEquivWitness}$ itself, or an equivalent dependent
  Sigma package with the projections below. It must not be merely
  $\operatorname{Nonempty}(\operatorname{OVBichromaticEquivWitness})$, and the
  two main fields must not have only the abstract type
  $\operatorname{FGReduction}$. The package contains two concrete subpackages
  \[
    R_{\mathrm{mono}\leftarrow\mathrm{bi}}
      \quad\text{and}\quad
    R_{\mathrm{bi}\leftarrow\mathrm{mono}},
  \]
  with the following mandatory fields.

  The source regimes are explicit large-parameter regimes, not arbitrary
  nonempty predicates:
  \[
    R_{\mathrm{mono}\leftarrow\mathrm{bi}}.\operatorname{SrcRegime}(n,\ell)
      \Longrightarrow n\ge2,\ \ell\ge1,
    \qquad
    R_{\mathrm{bi}\leftarrow\mathrm{mono}}.\operatorname{SrcRegime}(n,\ell)
      \Longrightarrow n\ge1,\ \ell\ge0,
  \]
  and each regime contains all sufficiently large parameter pairs in the
  displayed slice. In particular, neither source regime may be empty, `False',
  or a proposition with no inhabitants.

  The mono-to-bi subpackage has concrete fields:
  \[
    \operatorname{hasZero}(V)
      \Longleftrightarrow \exists i,\ v_i=0^\ell,
    \qquad
    \operatorname{toBi}(V)=(V,V)
  \]
  for every monochromatic instance $V=(v_1,\ldots,v_n)$. Its transformed
  algorithm, given a bichromatic OV algorithm $\mathcal A$, is definitionally:
  first return true if $\operatorname{hasZero}(V)$, and otherwise run
  $\mathcal A$ on $\operatorname{toBi}(V)=(V,V)$. The correctness field is the
  concrete equivalence
  \[
    \operatorname{Yes}_{\OV}(V)
    \Longleftrightarrow
    \operatorname{hasZero}(V)\ \vee\
    \operatorname{Yes}_{\mathrm{BOV}}(V,V),
  \]
  together with the proof that when $\operatorname{hasZero}(V)$ is false, every
  diagonal pair has nonzero inner product:
  $\inner{v_i}{v_i}\ne0$ for all $i$.

  The bi-to-mono subpackage has concrete parameter and instance maps
  \[
    (n,\ell)\mapsto(2n,\ell+2),\qquad
    a_i\mapsto a_i'=(a_i,1,0),\qquad
    b_j\mapsto b_j'=(b_j,0,1),
  \]
  and maps $(A,B)$ to the monochromatic list
  $A'\cup B'=(a_1',\ldots,a_n',b_1',\ldots,b_n')$. Its correctness field is the
  pair of displayed equalities and the resulting yes-predicate equivalence:
  \[
    \inner{a_i'}{a_{i'}'}\ge1,\qquad
    \inner{b_j'}{b_{j'}'}\ge1,\qquad
    \inner{a_i'}{b_j'}=\inner{a_i}{b_j},
  \]
  hence
  \[
    \operatorname{Yes}_{\mathrm{BOV}}(A,B)
    \Longleftrightarrow
    \operatorname{Yes}_{\OV}(A'\cup B').
  \]
  The Boolean reconstruction field for this direction is exactly the Boolean
  answer returned by the monochromatic OV algorithm on $A'\cup B'$.

  Both subpackages include runtime-transform fields, not just correctness:
  \[
    T_{\mathrm{mono}\leftarrow\mathrm{bi}}(n,\ell)
      \le C\bigl(n\ell+T_{\mathrm{bi}}(n,\ell)\bigr),
    \qquad
    T_{\mathrm{bi}\leftarrow\mathrm{mono}}(n,\ell)
      \le C\,T_{\mathrm{mono}}(2n,\ell+2)+C'n\ell,
  \]
  which preserve a truly subquadratic exponent up to polynomial factors in
  $\ell$. These fields must be propositions with the displayed inequalities as
  their conclusions. They may not have type `True', `Unit',
  `Nonempty True', or any other vacuous placeholder, and they may not omit the
  target runtime functions $T_{\mathrm{bi}}$ and $T_{\mathrm{mono}}$.

  More explicitly, the audited witness package must contain named fields whose
  contents are at least:
  \[
  \begin{array}{ll}
    \operatorname{monoToBiCorrect}:&
      \operatorname{Yes}_{\OV}(V)\Longleftrightarrow
      \operatorname{hasZero}(V)\vee\operatorname{Yes}_{\mathrm{BOV}}(V,V),\\
    \operatorname{monoToBiRuntime}:&
      \forall T_{\mathrm{bi}},\
      T_{\mathrm{mono}\leftarrow\mathrm{bi}}(n,\ell)
      \le C(n\ell+T_{\mathrm{bi}}(n,\ell)),\\
    \operatorname{biToMonoLeftInner}:&
      \inner{a_i'}{a_{i'}'}=\inner{a_i}{a_{i'}}+1,\\
    \operatorname{biToMonoRightInner}:&
      \inner{b_j'}{b_{j'}'}=\inner{b_j}{b_{j'}}+1,\\
    \operatorname{biToMonoCrossInner}:&
      \inner{a_i'}{b_j'}=\inner{a_i}{b_j},\\
    \operatorname{biToMonoCorrect}:&
      \operatorname{Yes}_{\mathrm{BOV}}(A,B)
      \Longleftrightarrow
      \operatorname{Yes}_{\OV}(A'\cup B'),\\
    \operatorname{biToMonoRuntime}:&
      \forall T_{\mathrm{mono}},\
      T_{\mathrm{bi}\leftarrow\mathrm{mono}}(n,\ell)
      \le C\,T_{\mathrm{mono}}(2n,\ell+2)+C'n\ell .
  \end{array}
  \]
  The three bi-to-mono inner-product fields may be imported directly from
  Lemma~\ref{lem:ov-bichromatic-append-construction}, but they must still be
  present in, or projected from, the reduction witness audited at this label. A
  proof object containing only empty-regime reductions, only abstract
  $\operatorname{FGReduction}$ fields, only runtime fields equal to `True', or
  only the implication
  $\TSQAlg{\text{bichromatic }\OV}\leftrightarrow\TSQAlg{\OV}$, is too weak and
  does not satisfy this lemma.
\end{lemma}
\begin{proof}
  \uses{def:ov, def:bichromatic-ov, def:inner-prod, lem:ov-bichromatic-append-construction, def:fg-reduction-predicates, def:tsq-alg-exists}
  ($\Leftarrow$) Suppose $\mathcal A$ solves bichromatic $\OV_{n,\ell}$ in
  $\bigO(n^{2-\eps}\operatorname{poly}(\ell))$. Given an $\OV_{n,\ell}$ instance
  $V = \{v_1, \ldots, v_n\}$, first check (in time $\bigO(n\ell)$) whether any
  $v_i$ is the all-zero vector; if so, this is a yes instance. Otherwise set
  $A = B = V$ and run $\mathcal A$ on $A,B$. Since no vector is zero,
  $\inner{v_i}{v_i} \ne 0$ for all $i$, so a pair $i \ne j$ with
  $\inner{v_i}{v_j} = 0$ exists iff $\mathcal A$ reports an orthogonal pair
  $a \in A$, $b \in B$. Total time $\bigO(n^{2-\eps}\operatorname{poly}(\ell) +
  n\ell) = \bigO(n^{2-\eps}\operatorname{poly}(\ell))$.
  This gives the witness
  $R_{\mathrm{mono}\leftarrow\mathrm{bi}}$ on the nonempty source regime of all
  sufficiently large monochromatic parameters; the instance map is $V\mapsto
  (V,V)$ after the explicit zero-vector branch, the output reconstruction is
  the returned Boolean answer, and the displayed runtime bound supplies the
  algorithm-transform field of Definition~\ref{def:fg-reduction-predicates}.

  ($\Rightarrow$) Suppose $\mathcal A'$ solves $\OV_{n,\ell}$ in
  $\bigO(n^{2-\eps}\operatorname{poly}(\ell))$. Given a bichromatic instance
  $A = \{a_1, \ldots, a_n\}$, $B = \{b_1, \ldots, b_n\}$ with $a_i, b_j \in
  \{0,1\}^{\ell}$, build $a_i' = (a_i, 1, 0) \in \R^{\ell+2}$ and
  $b_j' = (b_j, 0, 1) \in \R^{\ell+2}$, and let $A' = \{a_1', \ldots, a_n'\}$,
  $B' = \{b_1', \ldots, b_n'\}$. Then $\inner{a_i'}{a_j'} \ge 1$ and
  $\inner{b_i'}{b_j'} \ge 1$ for all $i,j$ (the appended coordinates contribute
  $1$), while $\inner{a_i'}{b_j'} = \inner{a_i}{b_j}$. Hence a pair of orthogonal
  vectors in $A' \cup B'$ can only be of the form $a_i', b_j'$, i.e.\ an
  orthogonal bichromatic pair $a_i, b_j$. Running $\mathcal A'$ on $A' \cup B'$
  (of size $2n$ in dimension $\ell+2$) decides the bichromatic instance in time
  $\bigO((2n)^{2-\eps}\operatorname{poly}(\ell+2)) =
  \bigO(n^{2-\eps}\operatorname{poly}(\ell))$.
  This gives the witness
  $R_{\mathrm{bi}\leftarrow\mathrm{mono}}$ on the nonempty source regime of all
  sufficiently large bichromatic parameters; its parameter map is
  $(n,\ell)\mapsto(2n,\ell+2)$, its instance map is the append-coordinate
  construction above, its Boolean reconstruction is the monochromatic answer,
  and its correctness theorem is exactly the equivalence between an orthogonal
  pair in $A'\cup B'$ and an orthogonal bichromatic pair in $A\times B$.
\end{proof}

\begin{lemma}[Min-IP is at least as hard as OV]
  \label{lem:minip-harder-ov}
  \uses{def:minip, def:gamma-minip, def:ov, def:fg-reduction-predicates, def:no-tsq-alg}
  For any $\gamma \ge 1$,
  $\AtLeastHard{\MinIP_{n,\ell}}{\OV_{n,\ell}}$ and
  $\AtLeastHard{\gamma\text{-}\MinIP_{n,\ell}}{\OV_{n,\ell}}$. Consequently,
  assuming OVC, for any $\eps>0$ there is $c>0$ such that
  $\NoTSQAlg{\MinIP_{n,c\log n}}$.

  The declaration audited at this label must be the concrete reduction witness
  package $\operatorname{OVToMinIPWitness}$, not only abstract
  $\operatorname{FGReduction}$ fields or the lower-bound conclusion. It contains
  two subwitnesses, one for exact Min-IP and one for $\gamma$-Min-IP. These
  subwitnesses are named
  \[
    \operatorname{exactReduction}
    \quad\text{and}\quad
    \operatorname{gammaReduction},
  \]
  but despite these names their types are not
  $\operatorname{Nonempty}(\operatorname{FGReduction}(\OV,\MinIP))$ and not
  $\operatorname{FGReduction}$ with hidden internals. They are concrete
  structures whose projections are exactly the maps, reconstruction rule,
  zero-correctness theorems and runtime bounds listed below. A generated Lean
  declaration in which `exactReduction' or `gammaReduction' is an abstract
  reduction witness is rejected.

  For every
  $\OV_{n,\ell}$ instance $V=(v_1,\ldots,v_n)$ both subwitnesses use the
  identity instance map
  \[
    \operatorname{toMinIP}(V)=V
  \]
  and, given an oracle output pair $(i,j)$, reconstruct the Boolean OV answer by
  the concrete test
  \[
    \operatorname{answer}(V,(i,j)) =
    \mathrm{true}
    \Longleftrightarrow
    \inner{v_i}{v_j}=0 .
  \]
  The exact subwitness has correctness fields
  \[
    \operatorname{MinInner}(V)=0
    \Longleftrightarrow
    \operatorname{Yes}_{\OV}(V),
    \qquad
    \operatorname{ValidOut}_{\MinIP}(V,(i,j))
    \wedge \inner{v_i}{v_j}=0
    \Longleftrightarrow
    \operatorname{Yes}_{\OV}(V).
  \]
  Here
  \[
    \operatorname{MinInner}(V)
      :=
      \min_{\substack{p,q\\p\ne q}}\inner{v_p}{v_q}
  \]
  is a required field or transparent local definition of the witness package,
  not an opaque symbol.
  The approximate subwitness has the additional zero-approximation field
  \[
    \operatorname{ValidOut}_{\gamma\text{-}\MinIP}(V,(i,j))
    \wedge \gamma\ge1
    \Longrightarrow
    \bigl(\operatorname{MinInner}(V)=0
      \Longleftrightarrow \inner{v_i}{v_j}=0\bigr),
  \]
  which is proved from
  $0\le\operatorname{MinInner}(V)\le\inner{v_i}{v_j}\le
  \gamma\operatorname{MinInner}(V)$. Both subwitnesses include the runtime
  field
  \[
    T_{\OV\leftarrow\MinIP}(n,\ell)
      \le C\bigl(T_{\MinIP}(n,\ell)+n\ell\bigr)
  \]
  and the analogous bound for $\gamma$-Min-IP,
  \[
    T_{\OV\leftarrow\gamma\MinIP}(n,\ell,\gamma)
      \le C\bigl(T_{\gamma\MinIP}(n,\ell,\gamma)+n\ell\bigr).
  \]
  These runtime fields mention the target algorithm runtime function and the
  $O(n\ell)$ cost of testing the returned pair. They are not tautological
  inequalities such as
  $T_{\MinIP}(n,\ell)+n\ell\le T_{\MinIP}(n,\ell)+n\ell$; their left-hand side
  is the runtime of the constructed OV algorithm, defined by running the oracle
  once and then computing $\inner{v_i}{v_j}$ for the returned pair. A proof
  object containing only
  $\operatorname{Nonempty}(\operatorname{FGReduction})$, an opaque algorithm
  transfer implication, or an OVC lower-bound proposition is too weak.
\end{lemma}
\begin{proof}
  \uses{def:minip, def:gamma-minip, def:ov, conjecture:ovc}
  Given an $\OV_{n,\ell}$ instance, run the (approximate) Min-IP algorithm to
  obtain a pair achieving the minimum inner product (or a $\gamma$-approximation
  of it) and test whether that value is $0$. There is an orthogonal pair iff the
  minimum inner product is $0$; and any multiplicative $\gamma$-approximation of
  $0$ must itself be $0$, so the $\gamma$-approximate version detects it too.
  Hence a truly subquadratic algorithm for ($\gamma$-)Min-IP gives one for OV,
  which is impossible under OVC (Conjecture~\ref{conjecture:ovc}).
\end{proof}

\begin{lemma}[Bichromatic Min-IP reduction, Lemma A.8]
  \label{lem:minip-bichromatic-reduction}
  \uses{def:minip, def:gamma-minip, def:minip-dec, def:bichromatic-minip, def:tsq-alg-exists}
  If $\TSQAlg{\text{bichromatic }\MinIP_{n,\ell}}$ exists with polynomial factors
  in $\ell$, then $\TSQAlg{\MinIP_{n,\ell}}$ exists with polynomial factors in
  $\ell$ (possibly with a different positive exponent). The same implication
  holds for $\gamma\text{-}\MinIP_{n,\ell}$ and for the decision version
  $\MinIP_{n,\ell,t}$.

  The audited declaration at this label must expose the recursive halving
  algorithm-transfer package
  $\operatorname{MinIPBichromaticRecursiveWitness}$. This package is the
  declaration audited at the label; the theorem statement is not merely the
  three implications between $\TSQAlg$ propositions. Its fields are:
  \[
    \operatorname{split}(V)=(V_L,V_R),\qquad
    |V_L|=\lceil |V|/2\rceil,\quad |V_R|=\lfloor |V|/2\rfloor,
  \]
  a call tree whose nodes are subsets $U\subseteq V$ and whose non-leaf node
  with split $(U_L,U_R)$ makes exactly the bichromatic oracle call
  $(U_L,U_R)$, and a reconstruction procedure taking all oracle answers and
  returning the best among the cross-split answers and the two recursively
  reconstructed within-half answers. The exact-search correctness field states
  \[
    \min_{\substack{p,q\in U\\p\ne q}}\inner{p}{q}
    =
    \min\!\left(
      \min_{\substack{p\in U_L\\q\in U_R}}\inner{p}{q},
      \min_{\substack{p,q\in U_L\\p\ne q}}\inner{p}{q},
      \min_{\substack{p,q\in U_R\\p\ne q}}\inner{p}{q}
    \right)
  \]
  at every recursive node, together with the theorem that the reconstructed
  pair realizes the global minimum of the original instance. The
  $\gamma$-approximate field states that if every oracle call returns a
  $\gamma$-approximate minimum for its cross part, then the best reconstructed
  candidate is a $\gamma$-approximate minimum for $V$. The threshold-decision
  field states that
  \[
    \exists p\ne q\in U,\ \inner{p}{q}\le t
  \]
  is equivalent to the disjunction of the corresponding cross, left-recursive,
  and right-recursive predicates.

  More explicitly, the package has named projections
  \[
  \begin{array}{ll}
    \operatorname{callTree}:&
      \text{the finite binary recursion tree of subsets }U\subseteq V,\\
    \operatorname{callsAtNode}:&
      U\mapsto\{(U_L,U_R)\}\text{ for every non-leaf node},\\
    \operatorname{reconstructExact}:&
      \text{selects the candidate with smallest inner product from the
      cross, left and right answers},\\
    \operatorname{exactNodeCorrectness}:&
      \text{the displayed minimum-decomposition equality for every node }U,\\
    \operatorname{exactGlobalCorrectness}:&
      \operatorname{ValidOut}_{\MinIP}(V,
        \operatorname{reconstructExact}(V,\operatorname{answers})),\\
    \operatorname{reconstructGamma}:&
      \text{the same smallest-score selection among approximate candidates},\\
    \operatorname{gammaNodeCorrectness}:&
      m_U\le s_U(\operatorname{reconstructGamma})\le\gamma m_U,\\
    \operatorname{thresholdNodeCorrectness}:&
      \operatorname{Yes}_{\MinIP_t}(U)\Longleftrightarrow
      \operatorname{Yes}_{\mathrm{cross},t}(U_L,U_R)\vee
      \operatorname{Yes}_{\MinIP_t}(U_L)\vee
      \operatorname{Yes}_{\MinIP_t}(U_R).
  \end{array}
  \]
  The last three fields are propositions containing the displayed mathematical
  relations. They may not have type `True', `Unit', `True -> True',
  `Nonempty True', or any other placeholder, and they may not be proved only
  from an abstract transfer implication.

  The runtime field is tied to the call tree:
  \[
    T_{\mathrm{mono}}(n,\ell)
      \le C n\ell+
      C\sum_{r=0}^{\lceil\log_2 n\rceil-1}
        2^r\,T_{\mathrm{bi}}\!\left(\left\lceil n/2^{r+1}\right\rceil,\ell\right),
  \]
  and the witness proves that if
  $T_{\mathrm{bi}}(m,\ell)\le C' m^{2-\eps}\operatorname{poly}(\ell)$, then the
  right-hand side is at most
  $C'' n^{2-\eps'}\operatorname{poly}(\ell)$ for some $\eps'>0$. A declaration
  whose fields are only implications between
  $\operatorname{Nonempty}(\operatorname{TSQAlgorithm})$ propositions, or which
  does not contain the split tree, optimum-in-some-call correctness, and the
  displayed runtime sum, is not this lemma.
\end{lemma}
\begin{proof}
  \uses{def:minip, def:bichromatic-minip}
  Given a $\MinIP_{n,\ell}$ instance $V = \{v_1, \ldots, v_n\}$, partition $V$
  into two halves $V_1, V_2$ of equal size and run $\mathcal A$ on the
  bichromatic instance $(V_1, V_2)$; this returns the minimal pair with one
  element in each half. Then recurse: split $V_1$ into two equal halves and find
  the minimal pair inside $V_1 \times V_1$, and likewise for $V_2$. The minimal
  pair of the whole instance lies either across the top-level split or inside one
  half, so it is found by some recursive call. The running time satisfies
  \[
    \operatorname{poly}(\ell)\cdot \sum_{i=1}^{\log n}
      \bigO\!\left(\left(\tfrac{n}{2^{i}}\right)^{2-\eps}\right)
    = \bigO\!\left(n^{2-\eps}\operatorname{poly}(\ell)\log n\right)
    \le \bigO\!\left(n^{2-\eps'}\operatorname{poly}(\ell)\right)
  \]
  for some $\eps'>0$ (the geometric sum is dominated by its first term). The same
  argument applies to $\gamma\text{-}\MinIP$ (the optimal pair still appears in
  some recursive call) and to $\MinIP_{n,\ell,t}$ (we examine all pairs across
  some call).
\end{proof}

\begin{lemma}[Bichromatic Max-IP reduction, Lemma A.12]
  \label{lem:maxip-bichromatic-reduction}
  \uses{def:maxip, def:gamma-maxip, def:bichromatic-maxip, def:tsq-alg-exists}
  If $\TSQAlg{\text{bichromatic }(\gamma\text{-})\MaxIP_{n,\ell}}$ exists with
  polynomial factors in $\ell$, then
  $\TSQAlg{(\gamma\text{-})\MaxIP_{n,\ell}}$ exists with polynomial factors in
  $\ell$ (possibly with a different positive exponent).

  The declaration audited at this label must expose its own recursive
  halving witness package
  $\operatorname{MaxIPBichromaticRecursiveWitness}$. It may reuse the same
  split-tree combinator as Lemma~\ref{lem:minip-bichromatic-reduction}, but it
  must instantiate it with maxima through its own concrete fields, not by
  pointing to the Min-IP recurrence as an abstract field. At every node
  $U=U_L\cup U_R$ its correctness theorem is
  \[
    \max_{\substack{p,q\in U\\p\ne q}}\inner{p}{q}
    =
    \max\!\left(
      \max_{\substack{p\in U_L\\q\in U_R}}\inner{p}{q},
      \max_{\substack{p,q\in U_L\\p\ne q}}\inner{p}{q},
      \max_{\substack{p,q\in U_R\\p\ne q}}\inner{p}{q}
    \right),
  \]
  and the reconstruction procedure returns the candidate with largest inner
  product among the cross-call and recursive candidates. For approximate
  Max-IP, if every bichromatic call returns a score in
  $[\gamma^{-1}M_{\mathrm{cross}},M_{\mathrm{cross}}]$, the reconstructed
  candidate must be proved to have score at least $\gamma^{-1}$ times the
  global maximum. The runtime field is the same call-tree sum as in
  Lemma~\ref{lem:minip-bichromatic-reduction}, with
  $T_{\mathrm{biMax}}$ in place of $T_{\mathrm{biMin}}$, and it must prove
  preservation of a positive truly-subquadratic exponent.

  The witness must have named projections
  \[
  \begin{array}{ll}
    \operatorname{callTree}:&
      \text{the finite split tree of subsets }U\subseteq V,\\
    \operatorname{reconstructExactMax}:&
      \text{selects the candidate with largest inner product},\\
    \operatorname{exactMaxNodeCorrectness}:&
      \text{the displayed maximum-decomposition equality for every node},\\
    \operatorname{exactMaxGlobalCorrectness}:&
      \operatorname{ValidOut}_{\MaxIP}(V,
        \operatorname{reconstructExactMax}(V,\operatorname{answers})),\\
    \operatorname{reconstructGammaMax}:&
      \text{selects the largest-score approximate candidate},\\
    \operatorname{gammaMaxNodeCorrectness}:&
      \gamma^{-1}M_U\le
      s_U(\operatorname{reconstructGammaMax})\le M_U,\\
    \operatorname{maxRuntimeSum}:&
      T_{\mathrm{monoMax}}(n,\ell)
      \le C n\ell+
      C\sum_{r=0}^{\lceil\log_2 n\rceil-1}
        2^r\,T_{\mathrm{biMax}}\!\left(
          \left\lceil n/2^{r+1}\right\rceil,\ell\right).
  \end{array}
  \]
  Fields named `exactNodeCorrectness', `gammaNodeCorrectness', or similar are
  acceptable only if their types are the maximum identities above. They may not
  be `True -> True' placeholders, and a proof object that only reuses a Min-IP
  recurrence without exposing the maximum reconstruction is not this lemma.
  Transfer implications alone do not satisfy this node.
\end{lemma}
\begin{proof}
  \uses{lem:minip-bichromatic-reduction, def:maxip, def:bichromatic-maxip}
  Identical to the proof of Lemma~\ref{lem:minip-bichromatic-reduction}: split
  the input recursively into halves and run the bichromatic Max-IP algorithm
  $\mathcal A$ on each pair of halves; the maximal pair appears in some recursive
  call, and the geometric sum of running times is dominated by the top term,
  giving $\bigO(n^{2-\eps'}\operatorname{poly}(\ell))$.
\end{proof}

\begin{lemma}[Max-IP is at least as hard as OV, Lemma A.13]
  \label{lem:maxip-harder-ov}
  \uses{def:maxip, def:bichromatic-maxip, def:ov, def:tsq-alg-exists, def:fg-reduction-predicates}
  $\AtLeastHard{\text{bichromatic }\MaxIP_{n,\ell}}{\OV_{n,\ell}}$: if
  $\TSQAlg{\text{bichromatic }\MaxIP_{n,\ell}}$ exists with polynomial factors in
  $\ell$, then $\TSQAlg{\OV_{n,\ell}}$ exists with polynomial factors in $\ell$.

  The declaration at this label must be the concrete witness package
  $\operatorname{OVToBichromaticMaxIPWitness}$. For an OV instance
  $V=(v_1,\ldots,v_n)$ it defines the weight classes
  \[
    S_r=\{v_i:\norm{v_i}_1=r\}\qquad (0\le r\le\ell)
  \]
  and for every ordered pair $(r,s)$ constructs the bichromatic Max-IP call
  \[
    \operatorname{call}_{r,s}(V)=\bigl(S_r,\overline{S_s}\bigr),
    \qquad
    \overline{u}[h]=1-u[h].
  \]
  The witness contains the inner-product identity
  \[
    v\in S_r,\ u\in S_s
    \Longrightarrow
    \inner{v}{\overline u}=r-\inner{v}{u},
  \]
  and the threshold equivalence
  \[
    \exists v\in S_r,\exists u\in S_s,\ \inner{v}{u}=0
    \Longleftrightarrow
    \max_{v\in S_r,\ w\in\overline{S_s}}\inner{v}{w}=r .
  \]
  Its reconstruction runs all $(\ell+1)^2$ calls, inspects the returned maximum
  score for each call, and outputs true iff some returned score is exactly the
  corresponding row weight $r$. The correctness field is the disjunction over
  all weight classes of the displayed equivalence, proving the reconstructed
  Boolean is true exactly on OV yes instances. The runtime field is
  \[
    T_{\OV\leftarrow\mathrm{BMaxIP}}(n,\ell)
      \le C n\ell+
        C\sum_{r=0}^{\ell}\sum_{s=0}^{\ell}
          T_{\mathrm{BMaxIP}}(|S_r|,|S_s|,\ell)
      \le C'(\ell+1)^2 T_{\mathrm{BMaxIP}}(n,\ell)+C''n\ell,
  \]
  which preserves truly subquadratic time up to polynomial factors in $\ell$.

  In field form, the audited witness contains at least
  \[
  \begin{array}{ll}
    \operatorname{weight}(v):& \norm{v}_1,\\
    \operatorname{weightClass}(V,r):& S_r=\{v_i:\norm{v_i}_1=r\},\\
    \operatorname{complementVector}(u):&
      \overline u\text{ with }\overline u[h]=1-u[h],\\
    \operatorname{callFamily}(V):&
      \{(S_r,\overline{S_s}) : 0\le r,s\le\ell\},\\
    \operatorname{callScore}(r,s,o):&
      \inner{v_o}{\overline u_o}\text{ for the returned cross pair }o,\\
    \operatorname{reconstruct}(V,\operatorname{answers}):&
      \mathrm{true}\Longleftrightarrow
      \exists r,s,\ \operatorname{callScore}(r,s,\operatorname{answers}_{r,s})=r,\\
    \operatorname{complementInner}:&
      v\in S_r,\ u\in S_s\Rightarrow
      \inner{v}{\overline u}=r-\inner{v}{u},\\
    \operatorname{weightClassCorrect}:&
      \operatorname{Yes}_{\OV}(V)\Longleftrightarrow
      \exists r,s,\ \max_{v\in S_r,w\in\overline{S_s}}\inner{v}{w}=r,\\
    \operatorname{runtimeWeightCalls}:&
      \text{the displayed double-sum runtime bound.}
  \end{array}
  \]
  None of these fields may be replaced by a single
  $\operatorname{FGReduction}$ field, a $\TSQAlg$ implication, or an abstract
  algorithm-transfer theorem. The returned score must be compared to the
  concrete weight $r$ of the left class, not merely to an unspecified threshold.
  A witness that is only
  $\operatorname{Nonempty}(\operatorname{FGReduction})$ or only an opaque
  implication from bichromatic Max-IP algorithms to OV algorithms omits the
  construction required here.
\end{lemma}
\begin{proof}
  \uses{def:maxip, def:bichromatic-maxip, def:ov, def:inner-prod}
  Given an $\OV_{n,\ell}$ instance $v_1, \ldots, v_n \in \{0,1\}^{\ell}$,
  partition the vectors into subsets $S_1, \ldots, S_{\ell}$ where $S_j$ contains
  all vectors with exactly $j$ ones. For each pair $1 \le i,j \le \ell$ let
  $\bar S_j$ consist of the vectors in $S_j$ with all entries flipped. Then for
  any $v \in S_i$ and $w \in \bar S_j$ (so $w = \bar u$ with $u \in S_j$), we have
  $\inner{v}{w} = i - \inner{v}{u}$, hence $\inner{v}{\bar u} = i$ iff
  $\inner{v}{u} = 0$. Therefore running the bichromatic Max-IP algorithm
  $\mathcal A$ on $(S_i, \bar S_j)$ and testing whether the maximum inner product
  equals $i$ reveals whether there is an orthogonal pair with one vector in $S_i$
  and one in $S_j$. Doing this for all $\ell^2$ pairs decides OV. The total time
  is $\bigO(\ell^{2}\cdot n^{2-\eps}\operatorname{poly}(\ell)) =
  \bigO(n^{2-\eps}\operatorname{poly}(\ell))$.
\end{proof}

\chapter{Hardness of document similarity}

\begin{theorem}[Hardness of $\gamma$-LSD, Theorem 3.1]
  \label{thm:gamma-lsd-hardness}
  \uses{def:gamma-lsd, def:ov, def:cosine, def:inner-prod, assumption:seth, conjecture:ovc, def:no-tsq-alg, def:log-parameter-regimes}
  Assuming SETH or OVC, for every $\eps>0$ there exists a constant $c>0$ such
  that $\NoTSQAlg{\gamma\text{-}\LSD_{n,\ell}}$ for every $\gamma \ge 1$ when
  $\ell = \logDim(c,n)$.

  The declaration audited at this label must expose the reduction package
  $\operatorname{OVToGammaLSDHardnessWitness}$ used to derive the lower bound.
  This package is the theorem's main data. It is not a field named
  `lowerBound', not a proposition $\NoTSQAlg{\gamma\text{-}\LSD}$ by itself,
  and not $\operatorname{Nonempty}(\operatorname{FGReduction})$ with hidden
  internals. A lower-bound proposition may be derived from the package only
  after the concrete maps, correctness proofs and runtime transfer below have
  been provided.
  Its source slice is the nonempty logarithmic regime
  $\ell=\logDim(c,n)$ for all sufficiently large $n$, and its target parameter
  is the same $(n,\ell,\gamma)$ with $\gamma\ge1$. For an OV instance
  $V=(v_1,\ldots,v_n)$ it defines
  \[
    \operatorname{hasZero}(V)
      \Longleftrightarrow \exists i,\ v_i=0^\ell,
    \qquad
    \operatorname{toLSD}(V)=
      \begin{cases}
        \text{unused} & \text{if }\operatorname{hasZero}(V),\\
        V & \text{otherwise,}
      \end{cases}
  \]
  so the $\gamma$-LSD oracle is called only on the valid nonzero-vector
  instance $V$ when no zero vector is present. Given an oracle output pair
  $(i^*,j^*)$, the reconstructed OV answer is
  \[
    \mathrm{true}
    \Longleftrightarrow
    \operatorname{hasZero}(V)\ \vee\
    \inner{v_{i^*}}{v_{j^*}}=0 .
  \]
  The witness includes an explicit oracle-call field
  \[
    \operatorname{calls}(V)=
      \begin{cases}
        \varnothing & \text{if }\operatorname{hasZero}(V),\\
        \{\,V\,\} & \text{otherwise, viewed as a valid }
          \gamma\text{-}\LSD_{n,\ell}\text{ instance,}
      \end{cases}
  \]
  and an answer-reconstruction field
  \[
    \operatorname{reconstruct}(V,\operatorname{answers})
      =
      \begin{cases}
        \mathrm{true} & \text{if }\operatorname{hasZero}(V),\\
        \bigl(\inner{v_{i^*}}{v_{j^*}}=0\bigr)
          & \text{for the returned LSD pair }(i^*,j^*)\text{ otherwise.}
      \end{cases}
  \]
  Thus a zero-vector OV instance is answered before the LSD oracle is called,
  and a nonzero instance is sent unchanged to the concrete nonzero-binary input
  relation of Definition~\ref{def:gamma-lsd}.
  The package contains the zero-cosine equivalence
  \[
    v_i\ne0,\ v_j\ne0
    \Longrightarrow
    \bigl(\cossim{v_i}{v_j}=0
      \Longleftrightarrow \inner{v_i}{v_j}=0\bigr),
  \]
  the minimum-zero equivalence
  \[
    \min_{p\ne q}\cossim{v_p}{v_q}=0
    \Longleftrightarrow
    \operatorname{Yes}_{\OV}(V)
  \]
  under $\neg\operatorname{hasZero}(V)$, and the approximation-zero field
  \[
    \operatorname{ValidOut}_{\gamma\text{-}\LSD}(V,(i^*,j^*))
    \wedge \gamma\ge1
    \Longrightarrow
    \left(
      \min_{p\ne q}\cossim{v_p}{v_q}=0
      \Longleftrightarrow
      \cossim{v_{i^*}}{v_{j^*}}=0
    \right).
  \]
  The last field must unfold the $\gamma$-LSD validity relation:
  if
  \[
    m=\min_{p\ne q}\cossim{v_p}{v_q}
    \quad\text{and}\quad
    m\le c^*\le\gamma m,\qquad c^*=\cossim{v_{i^*}}{v_{j^*}},
  \]
  then, using $m\ge0$ for nonzero binary vectors and $\gamma\ge1$,
  \[
    m=0 \Longleftrightarrow c^*=0.
  \]
  This is the multiplicative-zero argument; it cannot be replaced by a generic
  approximation theorem or by an abstract reduction correctness field.
  Its runtime field is
  \[
    T_{\OV\leftarrow\gamma\LSD}(n,\ell)
      \le C\bigl(n\ell+T_{\gamma\LSD}(n,\ell,\gamma)\bigr),
  \]
  and it proves this bound preserves the same positive truly-subquadratic
  exponent on the logarithmic dimension slice. The runtime field's left side is
  the runtime of the constructed OV algorithm with the zero-vector scan, at most
  one LSD oracle call, and the final inner-product test. It is not the
  tautology $n\ell+T_{\gamma\LSD}\le n\ell+T_{\gamma\LSD}$ and it is not a
  stand-alone asymptotic lower-bound proposition. A declaration that states only
  the high-level lower-bound proposition, without these maps, equivalences,
  reconstructed answer and runtime accounting, is not this theorem.
\end{theorem}
\begin{proof}
  \uses{def:gamma-lsd, def:ov, def:cosine, def:inner-prod, conjecture:ovc, thm:seth-implies-ovc, def:log-parameter-regimes}
  Assume for contradiction that some algorithm $\mathcal A$ solves
  $\gamma\text{-}\LSD_{n, \logDim(c,n)}$ in time $\bigO(n^{2-\eps})$ for some
  $\gamma \ge 1$, $\eps>0$ and \emph{every} constant $c>0$. We show that
  $\OV_{n, \logDim(c,n)}$ can then be solved in $\bigO(n^{2-\eps})$ time for
  every $c$, refuting OVC (and hence SETH, by
  Theorem~\ref{thm:seth-implies-ovc}).

  Given $v_1, \ldots, v_n \in \{0,1\}^{\ell}$ with
  $\ell = \logDim(c,n)$: if any vector is the zero vector (checkable in
  $\bigO(n\ell)$ time) output "yes". Otherwise run $\mathcal A$ to obtain
  $i^{*}, j^{*}$ with
  \[
    \min_{1 \le i,j \le n} \cossim{v_i}{v_j}
    \le \cossim{v_{i^{*}}}{v_{j^{*}}}
    \le \gamma \cdot \min_{1 \le i,j \le n} \cossim{v_i}{v_j}.
  \]
  Since all $\norm{v_i} > 0$, the quantity $\min_{i,j}\cossim{v_i}{v_j}$ equals
  $0$ iff there is an orthogonal pair (cosine similarity $0 \Leftrightarrow$
  inner product $0$). As $\gamma \ge 1$, $\cossim{v_{i^{*}}}{v_{j^{*}}} = 0$ iff
  the true minimum is $0$, i.e.\ iff $\mathcal A$ returns an orthogonal pair.
  Thus checking whether $\inner{v_{i^{*}}}{v_{j^{*}}} = 0$ decides
  $\OV_{n,\ell}$. The total time is $\bigO(n\ell + n^{2-\eps}) =
  \bigO(n^{2-\eps})$, refuting SETH/OVC.
\end{proof}

\begin{theorem}[Hardness of LSD and its variants, Corollary 3.2]
  \label{cor:lsd-hardness}
  \uses{def:lsd, def:gamma-lsd, def:bichromatic-gamma-lsd, def:lsd-dec, conjecture:ovc, assumption:seth, def:no-tsq-alg, def:log-parameter-regimes}
  Assuming SETH or OVC, for every $\eps>0$ there exists a constant $c>0$ such
  that $\NoTSQAlg{\LSD_{n,\ell}}$ when $\ell \ge \logDim(c,n)$. Moreover, the
  same lower bound holds for bichromatic
  $\gamma\text{-}\LSD_{n,\ell}$ for all $\gamma \ge 1$, and for $\LSD_{n,\ell,t}$
  for some $t \in [0,1]$.

  The declaration audited at this label must be the corollary witness package
  $\operatorname{LSDHardnessCorollaryWitness}$ with four non-vacuous fields:
  \begin{itemize}
    \item \emph{exact-search transfer}: an algorithm for exact $\LSD_{n,\ell}$
      yields a $\gamma$-LSD algorithm with $\gamma=1$ by running the exact
      search algorithm and returning the same pair, with the proof that exact
      minimizers satisfy the $\gamma=1$ approximation relation;
    \item \emph{larger-dimension transfer}: for every $\ell'\ge\logDim(c,n)$,
      the embedding $v\mapsto(v,0,\ldots,0)\in\{0,1\}^{\ell'}$ preserves
      nonzeroness and all pairwise cosine similarities, so a solver in dimension
      $\ell'$ gives one on the logarithmic hard slice;
    \item \emph{bichromatic transfer}: a solver for bichromatic
      $\gamma$-LSD on two color classes yields a monochromatic $\gamma$-LSD
      solver by the standard recursive split construction over a single list,
      with the same split-tree, reconstruction and runtime accounting as
      Lemma~\ref{lem:minip-bichromatic-reduction}, but using cross-color cosine
      minima instead of inner-product minima;
    \item \emph{threshold transfer}: if every threshold family
      $\LSD_{n,\ell,t}$ for $t\in[0,1]$ had a truly subquadratic algorithm, then
      exact $\LSD_{n,\ell}$ would be solved by enumerating or binary searching
      the finite set
      \[
        \mathcal T_{\ell}
        =
        \left\{\frac{s}{\sqrt{ab}}:
          0\le s\le\ell,\ 1\le a,b\le\ell\right\}
      \]
      of all possible cosine values of nonzero binary vectors and returning a
      pair at the least feasible threshold. The package must therefore prove the
      existential conclusion
      \[
        \exists t\in[0,1],\quad
        \NoTSQAlg{\LSD_{n,\ell,t}}
      \]
      on the logarithmic slice; it must not state a lower bound for an
      unspecified or all-purpose `def_lsd_dec' family without binding the
      witness threshold $t$.
  \end{itemize}
  The four fields have the following required concrete shapes:
  \[
  \begin{array}{ll}
    \operatorname{exactToGammaOne}:&
      \operatorname{ValidOut}_{\LSD}(V,(i,j))
      \Rightarrow
      \operatorname{ValidOut}_{1\text{-}\LSD}(V,(i,j)),\\
    \operatorname{padToLargerDim}:&
      v\mapsto(v,0^{\ell'-\ell})\text{ with }
      \cossim{(v,0)}{(w,0)}=\cossim{v}{w},\\
    \operatorname{largerDimRuntime}:&
      T_{\LSD,\ell}(n)\le C(n\ell'+T_{\LSD,\ell'}(n)),\\
    \operatorname{bichromaticSplitTree}:&
      \text{the recursive halving tree over one list }V,\\
    \operatorname{bichromaticCosineNodeCorrectness}:&
      \min_{p\ne q\in U}\cossim{p}{q}
      =
      \min\!\left(
        \min_{p\in U_L,q\in U_R}\cossim{p}{q},
        \min_{p\ne q\in U_L}\cossim{p}{q},
        \min_{p\ne q\in U_R}\cossim{p}{q}
      \right),\\
    \operatorname{bichromaticGammaCorrectness}:&
      \text{best reconstructed cross/recursive candidate is a }
      \gamma\text{-approximate cosine minimum},\\
    \operatorname{thresholdValues}:&
      \mathcal T_\ell=
      \{s/\sqrt{ab}:0\le s\le\ell,\ 1\le a,b\le\ell\},\\
    \operatorname{thresholdSearchCorrectness}:&
      \text{enumerating }\mathcal T_\ell\text{ recovers the least cosine
      value and a minimizing pair},\\
    \operatorname{existsHardThreshold}:&
      \exists t,\ 0\le t\le1\ \wedge\
      \NoTSQAlg{\LSD_{n,\ell,t}}.
  \end{array}
  \]
  None of these fields may have type `True', `Unit', `True -> True', or be only
  an implication between abstract $\operatorname{TSQAlgorithm}$ existence
  propositions. In particular, the threshold field must quantify and return a
  threshold-specific family $\LSD_{n,\ell,t}$ with the witness $t$ and proof
  $0\le t\le1$; the conclusion
  $\NoTSQAlg{\text{def\_lsd\_dec}}$ without a bound threshold is not this
  corollary.
  These fields must retain the parameter regime
  $\ell\ge\logDim(c,n)$ and the bichromatic $\gamma$-LSD claim. A corollary
  statement that only mentions the abstract families
  $\LSD$, $\gamma$-LSD and $\LSD$-decision, or omits the existential threshold,
  is weaker than the paper's Corollary 3.2.
\end{theorem}
\begin{proof}
  \uses{thm:gamma-lsd-hardness, def:lsd, def:lsd-dec, def:no-tsq-alg, def:log-parameter-regimes}
  Since $\gamma\text{-}\LSD_{n,\ell}$ (with $\gamma = 1$) and bichromatic
  $\gamma\text{-}\LSD_{n,\ell}$ are both easier than $\LSD_{n,\ell}$, the lower
  bound of Theorem~\ref{thm:gamma-lsd-hardness} transfers to them. For the
  decision version: there must exist $t \in [0,1]$ such that $\LSD_{n,\ell,t}$
  cannot be solved in $\bigO(n^{2-\eps})$ time when $\ell = \logDim(c,n)$,
  because otherwise binary search over the $\bigO(\ell^{3})$ possible threshold
  values would give a $\bigO(n^{2-\eps})$ algorithm for $\LSD_{n,\ell}$,
  contradicting the result for $\LSD_{n,\ell}$. (Making $\ell$ larger only makes
  the problem harder, giving the bound for all $\ell \ge \logDim(c,n)$.)
\end{proof}

\begin{theorem}[Hardness of $\gamma$-MSD, Theorem 3.3 / B.1]
  \label{thm:gamma-msd-hardness}
  \uses{def:gamma-msd, assumption:seth, conjecture:ovc, def:no-tsq-alg, def:log-parameter-regimes}
  Assuming SETH or OVC, for every $\eps>0$ there exists a constant $c>0$ such
  that $\NoTSQAlg{\gamma\text{-}\MSD_{n,\msdHardDim(c,n)}}$ whenever
  $\msdHardGammaAllowed(n,\gamma)$, i.e.
  \[
    \ell = \msdHardDim(c,n)
      = (\log n)^{\frac{c\log n}{(\log\log n)^{2}}}
    \quad\text{and}\quad
    \gamma \le \left(1 + \frac{1}{\log\log n}\right)^{\frac{\log n}{(\log\log n)^{2}}}
    = 2^{(\log n)^{1-o(1)}}.
  \]
\end{theorem}
\begin{proof}
  \uses{lem:msd-tensor-reduction, lem:msd-to-additive-msd, lem:additive-msd-maxip-equiv, thm:additive-bmaxip-to-ov, conjecture:ovc, thm:seth-implies-ovc, def:log-parameter-regimes}
  This follows from a combination of Lemma~\ref{lem:msd-tensor-reduction},
  Lemma~\ref{lem:msd-to-additive-msd},
  Lemma~\ref{lem:additive-msd-maxip-equiv} and
  Theorem~\ref{thm:additive-bmaxip-to-ov}. A truly subquadratic algorithm for
  $\gamma\text{-}\MSD_{n,\msdHardDim(c,n)}$ with
  $\msdHardGammaAllowed(n,\gamma)$ would, by
  Lemma~\ref{lem:msd-tensor-reduction}, give a truly subquadratic algorithm for
  $\mildMsdGamma(n)\text{-}\MSD_{n,\polylogDim(k,n)}$ for any constant $k>0$; by
  Lemma~\ref{lem:msd-to-additive-msd} (which itself combines
  Theorem~\ref{thm:additive-maxip-karthik} and
  Lemma~\ref{lem:additive-msd-maxip-equiv}) this gives a truly subquadratic
  algorithm for bichromatic
  $\frac{\logn(n)}{\logDim(c,n)}$-Additive-$\MSD_{n,\logDim(c,n)}$;
  and finally Theorem~\ref{thm:additive-bmaxip-to-ov} turns a truly subquadratic
  algorithm for the corresponding bichromatic additive Max-IP into a truly
  subquadratic algorithm for $\OV_{n, \logDim(c',n)}$, refuting SETH and OVC
  (Conjecture~\ref{conjecture:ovc}, Theorem~\ref{thm:seth-implies-ovc}). The
  hardness also transfers to harder problems including $\MSD_{n,\ell}$ and
  bichromatic $\gamma\text{-}\MSD_{n,\ell}$.
\end{proof}

\begin{theorem}[Hardness of MSD and its variants, Corollary 3.4]
  \label{cor:msd-hardness}
  \uses{def:msd, def:gamma-msd, def:msd-dec, assumption:seth, conjecture:ovc, def:no-tsq-alg, def:log-parameter-regimes}
  Assuming SETH or OVC, for every $\eps>0$ there exists a constant $c>0$ such
  that $\NoTSQAlg{\MSD_{n,\ell}}$ when
  $\ell \ge \msdHardDim(c,n)$. Moreover the same lower bound holds for
  bichromatic $\gamma\text{-}\MSD_{n,\ell}$ whenever
  $\msdHardGammaAllowed(n,\gamma)$ and for $\MSD_{n,\ell,t}$ for some
  $t \in [0,1]$.
\end{theorem}
\begin{proof}
  \uses{thm:gamma-msd-hardness, def:msd, def:msd-dec, def:no-tsq-alg, def:log-parameter-regimes}
  As in Corollary~\ref{cor:lsd-hardness}: $\gamma\text{-}\MSD$ and bichromatic
  $\gamma\text{-}\MSD$ are easier than $\MSD$, so the bound of
  Theorem~\ref{thm:gamma-msd-hardness} transfers; and for the decision version
  $\MSD_{n,\ell,t}$, the number of distinct relevant thresholds $t$ is
  $\bigO(\ell^{3})$, so a truly subquadratic algorithm for all $t$ would solve
  $\MSD_{n,\ell}$ by binary search, contradicting
  Theorem~\ref{thm:gamma-msd-hardness}; hence some $t$ is hard.
\end{proof}

\chapter{Proof of the $\gamma$-MSD hardness (Appendix B)}

\begin{definition}[Bichromatic $\gamma$-Additive-MSD]
  \label{def:additive-msd}
  \uses{def:cosine, def:log-parameter-regimes, def:parameter-sensitive-placeholder}
  For parameters $n,\ell\in\N$ and additive gap $\gamma\ge 0$, an instance of
  bichromatic $\gamma$-Additive-MSD consists of two $n$-tuples
  $A=(a_1,\ldots,a_n)$ and $B=(b_1,\ldots,b_n)$ of nonzero vectors in
  $\{0,1\}^{\ell}$ and a threshold $\alpha \in [0,1]$. Let
  $c_I(i,j)=\cossim{a_i}{b_j}$. This is a promise decision family whose two
  predicates are:
  \[
    \operatorname{Yes}_{\gamma}(I)
    \Longleftrightarrow
    \exists\,i,j,\ c_I(i,j)\ge\alpha,
  \]
  and
  \[
    \operatorname{No}_{\gamma}(I)
    \Longleftrightarrow
    \forall\,i,j,\ c_I(i,j)<\alpha-\gamma.
  \]
  A solver for this promise problem must output true on every yes instance and
  false on every no instance; inputs satisfying neither predicate lie in the
  promise gap and need not be classified. The additive gap $\gamma$ appears
  specifically in the no predicate as the separation $\alpha-\gamma$. Recording
  only the yes predicate, or treating $\gamma$ as unused metadata, omits half of
  the problem.
\end{definition}

\begin{definition}[Bichromatic $\gamma$-Additive-Max-IP]
  \label{def:additive-maxip}
  \uses{def:inner-prod, def:log-parameter-regimes, def:parameter-sensitive-placeholder}
  For parameters $n,\ell\in\N$ and additive gap $\gamma\ge 0$, an instance of
  bichromatic $\gamma$-Additive-Max-IP ($\gamma$-Additive-BMax-IP) consists of
  two $n$-tuples $A=(a_1,\ldots,a_n)$ and $B=(b_1,\ldots,b_n)$ in
  $\{0,1\}^{\ell}$ and a threshold $\alpha$. Let
  $s_I(i,j)=\inner{a_i}{b_j}$. This is a promise decision family whose two
  predicates are:
  \[
    \operatorname{Yes}_{\gamma}(I)
    \Longleftrightarrow
    \exists\,i,j,\ s_I(i,j)\ge\alpha,
  \]
  and
  \[
    \operatorname{No}_{\gamma}(I)
    \Longleftrightarrow
    \forall\,i,j,\ s_I(i,j)<\alpha-\gamma.
  \]
  A solver must output true on all yes instances and false on all no instances;
  inputs in the gap between $\alpha-\gamma$ and $\alpha$ are outside the promise.
  The formal problem must store both vector lists, the threshold $\alpha$, both
  predicates, and the additive gap $\gamma$. A declaration containing only the
  existential yes predicate is not the additive-gap problem used by Lemma B.6.
\end{definition}

\begin{lemma}[Tensoring amplifies the MSD approximation factor, Lemma B.2]
  \label{lem:msd-tensor-reduction}
  \uses{def:gamma-msd, def:kron, lem:kron-power-cosine, def:tsq-alg-exists, def:log-parameter-regimes}
  Suppose $\TSQAlg{\gamma\text{-}\MSD_{n,\ell}}$ holds with
  $\ell = \msdHardDim(c,n)$ for any constant $c>0$ and
  $\msdHardGammaAllowed(n,\gamma)$, running in $\bigO(n^{2-\eps})$ time for any
  $\eps>0$. Then
  $\TSQAlg{\mildMsdGamma(n)\text{-}\MSD_{n,\polylogDim(k,n)}}$ holds with the
  same truly subquadratic exponent for any constant $k>0$.
\end{lemma}
\begin{proof}
  \uses{lem:kron-power-cosine, def:gamma-msd, def:kron, def:cosine, def:log-parameter-regimes}
  Let $v_1, \ldots, v_n \in \{0,1\}^{\polylogDim(k,n)}$ be an instance of
  $\mildMsdGamma(n)\text{-}\MSD_{n,\polylogDim(k,n)}$. Construct
  $V' = \{v_1^{\otimes q}, \ldots, v_n^{\otimes q}\}$ for
  $q = \frac{\log n}{(\log\log n)^{2}}$. The vectors of $V'$ have dimension
  $(\log n)^{kq} = (\log n)^{\frac{k\log n}{(\log\log n)^{2}}}$, so the assumed
  algorithm applies and finds a $\gamma$-approximation of MSD on $V'$ for some
  $\gamma \le (1 + \tfrac{1}{\log\log n})^{q}$. By
  Lemma~\ref{lem:kron-power-cosine},
  \[
    \frac{\inner{v_i^{\otimes q}}{v_j^{\otimes q}}}
         {\norm{v_i^{\otimes q}}\cdot\norm{v_j^{\otimes q}}}
    = \left(\cossim{v_i}{v_j}\right)^{q}
  \]
  for all $i,j$. Therefore a $\gamma$-approximation of MSD on $V'$ is a
  $\gamma^{1/q} = \mildMsdGamma(n)$-approximation of MSD on
  $v_1, \ldots, v_n$. The running time is preserved up to the $\bigO(n^{2-\eps})$
  bound.
\end{proof}

\begin{assumption}[Karthik--Manurangsi reduction, Theorem B.4]
  \label{thm:additive-maxip-karthik}
  \uses{def:gamma-msd, def:additive-maxip, def:tsq-alg-exists, def:log-parameter-regimes}
  This is the cited Theorem 6.2 of Karthik--Manurangsi (2020), used by the paper
  without reproducing the graph-construction proof. Its content is:
  Suppose
  $\TSQAlg{\mildMsdGamma(n)\text{-}\MSD_{n,\polylogDim(k,n)}}$ holds for any
  constant $k>0$ and any truly subquadratic exponent. Then
  $\TSQAlg{\text{bichromatic }\logn(n)\text{-Additive-Max-IP}_{n,\logDim(c,n)}}$
  holds for any constant $c>0$ with some positive truly subquadratic exponent.
  A disposable Lean implementation should treat this node as an axiom/imported
  external theorem unless it formalizes the full graph construction and its
  approximation-gap analysis.
\end{assumption}

\begin{lemma}[Additive-MSD is equivalent to Additive-Max-IP, Lemma B.5]
  \label{lem:additive-msd-maxip-equiv}
  \uses{def:additive-msd, def:additive-maxip, def:fg-reduction-predicates, def:log-parameter-regimes}
  $\SubquadEquiv{\text{bichromatic }\frac{\logn(n)}{\ell}\text{-Additive-}\MSD_{n,\ell}}
  {\text{bichromatic }\logn(n)\text{-Additive-Max-IP}_{n,\ell'}}$ when
  $\ell = \bigO(\log n)$ and $\ell' = \bigO(\log n)$.
\end{lemma}
\begin{proof}
  \uses{def:additive-msd, def:additive-maxip, def:inner-prod, def:l2norm, def:log-parameter-regimes}
  ($\Leftarrow$) Given a bichromatic $\logn(n)$-Additive-Max-IP$_{n,\ell'}$
  instance with sets $A = \{a_1, \ldots, a_n\}$, $B = \{b_1, \ldots, b_n\}$ and
  threshold $\alpha$, construct $A' = \{a_1', \ldots, a_n'\}$, $B' = \{b_1',
  \ldots, b_n'\}$ as follows: for each $a_i$ first attach $\ell - \norm{a_i}_1$
  ones at the end and then another $\ell + \norm{a_i}_1$ zeros to obtain $a_i'
  \in \{0,1\}^{3\ell}$, and for each $b_j$ attach $\ell + \norm{b_j}_1$ zeros and
  then another $\ell - \norm{b_j}_1$ ones to obtain $b_j' \in \{0,1\}^{3\ell}$.
  Now every $a_i'$ and $b_j'$ has exactly $\ell$ ones, so $\norm{a_i'} =
  \norm{b_j'} = \sqrt{\ell}$, and $\inner{a_i'}{b_j'} = \inner{a_i}{b_j}$ by the
  construction (the appended coordinates never overlap on a $1$). Hence running an
  $\frac{\logn(n)}{\ell}$-Additive-MSD algorithm on $A', B'$ with threshold
  $\frac{\alpha}{\ell}$ distinguishes:
  \begin{itemize}
    \item there exists $(a',b')$ with $\frac{\inner{a'}{b'}}{\norm{a'}\norm{b'}}
      \ge \frac{\alpha}{\ell}$, equivalent to $\inner{a}{b} \ge \alpha$; or
    \item for all $(a',b')$, $\frac{\inner{a'}{b'}}{\norm{a'}\norm{b'}} <
      \frac{\alpha}{\ell} - \frac{\logn(n)}{\ell}$, equivalent to
      $\inner{a}{b} < \alpha - \logn(n)$ for all $a \in A$, $b \in B$.
  \end{itemize}
  The running time is $\bigO(n^{2-\eps}\operatorname{poly}(3\ell)) =
  \bigO(n^{2-\eps}\operatorname{poly}(\ell))$.

  ($\Rightarrow$) Conversely, given a bichromatic
  $\frac{\logn(n)}{\ell}$-Additive-MSD$_{n,\ell}$ instance with sets $A, B$ and
  threshold $\alpha$, build the exact same $A', B'$ so that
  $\inner{a_i'}{b_j'} = \inner{a_i}{b_j}$ and $\norm{a_i'} = \norm{b_j'} =
  \sqrt{\ell}$ for all $i,j$. The same argument shows that running an
  Additive-Max-IP algorithm on $A', B'$ with threshold $\alpha\cdot\ell$ solves
  the Additive-MSD instance.
\end{proof}

\begin{lemma}[Bichromatic Additive-MSD reduces to Additive-Max-IP, Lemma B.3]
  \label{lem:msd-to-additive-msd}
  \uses{def:gamma-msd, def:additive-msd, def:additive-maxip, def:tsq-alg-exists, def:log-parameter-regimes}
  If
  $\TSQAlg{\mildMsdGamma(n)\text{-}\MSD_{n,\polylogDim(k,n)}}$ holds for any
  constant $k>0$, then
  $\TSQAlg{\text{bichromatic }\frac{\logn(n)}{\logDim(c,n)}\text{-Additive-}
  \MSD_{n,\logDim(c,n)}}$ holds for any $c>0$ with some positive truly
  subquadratic exponent.
\end{lemma}
\begin{proof}
  \uses{thm:additive-maxip-karthik, lem:additive-msd-maxip-equiv, def:log-parameter-regimes}
  By Theorem~\ref{thm:additive-maxip-karthik}, the assumed
  $\mildMsdGamma(n)$-MSD algorithm yields a truly subquadratic algorithm for
  bichromatic $\logn(n)$-Additive-Max-IP. By
  Lemma~\ref{lem:additive-msd-maxip-equiv}, bichromatic $\logn(n)$-Additive-
  Max-IP is subquadratic equivalent to bichromatic
  $\frac{\logn(n)}{\ell}$-Additive-MSD, so the latter is also truly
  subquadratic.
\end{proof}

\begin{assumption}[Additive-BMax-IP refutes OVC, Lemma B.6 (Chen 2020)]
  \label{thm:additive-bmaxip-to-ov}
  \uses{def:additive-maxip, def:ov, conjecture:ovc, assumption:seth, def:tsq-alg-exists, def:log-parameter-regimes}
  This is a cited external dependency from Chen (2020), not a proof reproduced
  in this paper. Its content is:
  If $\TSQAlg{\logn(n)\text{-Additive-BMax-IP}_{n,\logDim(c,n)}}$ holds for any
  constant $c>0$ with some positive truly subquadratic exponent, then
  $\TSQAlg{\OV_{n,\logDim(c',n)}}$ holds for any constant $c'>0$ with some
  positive truly subquadratic exponent, refuting SETH and OVC.
  A formalization may register this as an axiom/imported theorem. It should not
  attempt to prove this node from the definitions alone unless it also
  formalizes Chen's additive maximum-inner-product hardness theorem.
\end{assumption}

\chapter{Representational strength of transformers}

\begin{theorem}[An attention unit solves OV, Theorem 4.1]
  \label{thm:attention-solves-ov}
  \uses{def:transformer, def:tf-solves, def:ov, def:attention-solves-predicate}
  An attention unit with input and output MLPs and parameters $d = \ell$,
  $d_{\mathrm{in}} = \ell$, $d_{\mathrm{out}} = 1$, $m \ge \ell + 1$ can solve
  $\OV_{n,\ell}$; in the notation of
  Definition~\ref{def:attention-solves-predicate}, $\AttentionSolves{\OV_{n,\ell}}$.
\end{theorem}
\begin{proof}
  \uses{def:attention, def:softmax, def:tf-solves, def:ov, lem:mlp-threshold, def:inner-prod}
  Let $v_1, \ldots, v_n \in \{0,1\}^{\ell}$ be an $\OV_{n,\ell}$ instance; define
  $v_{n+1} := 0^{\ell}$ and let $X \in \R^{(n+1)\times\ell}$ have rows
  $X_{i,:} = v_i$. Since $m \ge \ell + 1$, choose $Q, K \in \R^{\ell \times m}$
  whose score bilinear form satisfies $M_{Q,K} = -3\log n \cdot I_{\ell}$, let
  $V \in \R^{\ell\times 1}$ be the all-one matrix, and take $\varphi_1$ to be the
  identity. We claim that if there is a pair $i \ne j$ with
  $\inner{v_i}{v_j} = 0$ then $A_{Q,K,V}(X)$ has an entry $\ge \frac{1}{n+1}$,
  and otherwise all entries are $\le
  \frac{1}{(n+1)^{1.5}}$; the second MLP $\varphi_2$ then outputs $1$ iff some
  entry is $\ge \frac{1}{n+1}$ (existence guaranteed by
  Lemma~\ref{lem:mlp-threshold}).

  The $i$-th entry of $A_{Q,K,V}(X)$ equals
  \[
    \sum_{j=1}^{n} \frac{\exp(-3\log n \cdot \inner{v_i}{v_j})}
      {\sum_{k=1}^{n}\exp(-3\log n \cdot \inner{v_i}{v_k}) + 1}\cdot \norm{v_j}_1
    = \frac{\sum_{j=1}^{n} n^{-3\inner{v_i}{v_j}}\cdot \norm{v_j}_1}
           {\sum_{k=1}^{n} n^{-3\inner{v_i}{v_k}} + 1},
  \]
  where $\norm{v_j}_1 = (XV)_j$ (the all-one $V$ sums the entries of $v_j$) and
  the $+1$ in the denominator is the contribution of $v_{n+1} = 0^{\ell}$ (giving
  $n^{0} = 1$). If there is a pair with $\inner{v_{i^{*}}}{v_{j^{*}}} = 0$, the
  $i^{*}$-th entry is at least
  \[
    \frac{n^{-3\inner{v_{i^{*}}}{v_{j^{*}}}}\cdot \norm{v_{j^{*}}}_1}{n+1}
    \ge \frac{1}{n+1},
  \]
  since $\norm{v_{j^{*}}}_1 \ge 1$ and the denominator is at most $n+1$.
  Otherwise $\inner{v_i}{v_j} \ge 1$ for all $i,j$, so the $i$-th entry is at most
  \[
    \sum_{j=1}^{n} \frac{n^{-3\inner{v_i}{v_j}} \cdot \norm{v_j}_1}{1}
    \le \sum_{j=1}^{n} \frac{\ell}{n^{3\inner{v_i}{v_j}}}
    \le n\cdot \frac{\ell}{n^{3}} < \frac{1}{n^{1.5}},
  \]
  using $\norm{v_j}_1 \le \ell$ and $\ell = \bigO(\operatorname{polylog} n)$.
\end{proof}

\begin{theorem}[An attention unit solves Max-IP and Min-IP decision, Theorem C.1]
  \label{thm:attention-solves-maxip-minip}
  \uses{def:transformer, def:tf-solves, def:maxip-dec, def:minip-dec, def:attention-solves-predicate}
  An attention unit with input and output MLPs and parameters $d = \ell$,
  $d_{\mathrm{in}} = \ell + 1$, $d_{\mathrm{out}} = 1$, $m \ge \ell + 1$ can
  solve $\MaxIP_{n,\ell,t}$ and $\MinIP_{n,\ell,t}$ for $1 \le t \le \ell$; that
  is, $\AttentionSolves{\MaxIP_{n,\ell,t}}$ and
  $\AttentionSolves{\MinIP_{n,\ell,t}}$.
\end{theorem}
\begin{proof}
  \uses{def:attention, def:softmax, def:maxip-dec, def:minip-dec, lem:mlp-threshold, lem:mlp-append, def:inner-prod}
  \textbf{Max-IP.} Given $v_1, \ldots, v_n \in \{0,1\}^{\ell}$ and $V \in
  \R^{n\times\ell}$ with $V_{i,:} = v_i$, set $x_i = (v_i, 1) \in
  \{0,1\}^{\ell+1}$, $x_{n+1} = (0, \ldots, 0, t+1) \in \R^{\ell+1}$, and let
  $X \in \R^{(n+1)\times(\ell+1)}$ have rows $X_{i,:} = x_i$. Choose $Q, K \in
  \R^{(\ell+1)\times m}$ whose score bilinear form satisfies
  $M_{Q,K} = 3\log n \cdot I_{\ell+1}$, and
  $\mathcal V = (1, \ldots, 1, 0) \in \R^{\ell+1}$. The input MLP $\varphi_1$
  produces the rows $x_i$ from the $v_i$ via Lemma~\ref{lem:mlp-append}. Then the
  $i$-th entry of $A_{Q,K,\mathcal V}(X)$ equals
  \[
    \frac{\sum_{j=1}^{n} n^{3\inner{x_i}{x_j}}\cdot \norm{v_j}_1}
         {\sum_{k=1}^{n} n^{3\inner{x_i}{x_k}} + n^{3(t+1)}}.
  \]
  Let $i^{*} \ne j^{*}$ maximize $\inner{v_{i^{*}}}{v_{j^{*}}}$. If
  $\inner{v_{i^{*}}}{v_{j^{*}}} \ge t$, then $\inner{x_{i^{*}}}{x_{j^{*}}} \ge
  t+1$ and the $i^{*}$-th entry is at least
  \[
    \frac{n^{3\inner{x_{i^{*}}}{x_{j^{*}}}}\cdot \norm{v_{j^{*}}}_1}
         {\sum_{k} n^{3\inner{x_{i^{*}}}{x_k}} + n^{3(t+1)}}
    \ge \frac{\norm{v_{j^{*}}}_1}{n+1} \ge \frac{1}{n+1}.
  \]
  If instead $\inner{v_i}{v_j} \le t-1$ for all $i \ne j$, then
  $\inner{x_i}{x_j} \le t$ and the $i$-th entry is at most
  \[
    \frac{\sum_{j=1}^{n} n^{3\inner{x_i}{x_j}}\cdot \norm{v_j}_1}{n^{3(t+1)}}
    \le \frac{\ell\cdot n\cdot n^{3t}}{n^{3(t+1)}} = \frac{\ell}{n^{2}}
    < \frac{1}{(n+1)^{1.5}}.
  \]
  Thus $\varphi_2$ outputs $1$ iff some entry is $\ge \frac{1}{n+1}$
  (Lemma~\ref{lem:mlp-threshold}), solving $\MaxIP_{n,\ell,t}$.

  \textbf{Min-IP.} The proof is the same except the score bilinear form is
  $M_{Q,K} = -3\log n \cdot I_{\ell+1}$ and
  $x_{n+1} = (0, \ldots, 0, -(t+1)) \in \R^{\ell+1}$. The $i$-th entry of
  $A_{Q,K,\mathcal V}(X)$ becomes
  \[
    \frac{\sum_{j=1}^{n} n^{-3\inner{x_i}{x_j}}\cdot \norm{v_j}_1}
         {\sum_{k=1}^{n} n^{-3\inner{x_i}{x_k}} + n^{-3(t+1)}}.
  \]
  Let $i^{*}, j^{*}$ minimize $\inner{x_{i^{*}}}{x_{j^{*}}}$. If
  $\inner{v_{i^{*}}}{v_{j^{*}}} \le t$ then $\inner{x_{i^{*}}}{x_{j^{*}}} \le t+1$
  and the $i^{*}$-th entry is at least $\frac{1}{n+1}$; if $\inner{v_i}{v_j} \ge
  t+1$ for all $i \ne j$ then $\inner{x_i}{x_j} \ge t+2$ and the $i$-th entry is
  at most $\frac{\ell}{n^{2}} < \frac{1}{(n+1)^{1.5}}$. Again $\varphi_2$
  thresholds at $\frac{1}{n+1}$.
\end{proof}

\begin{theorem}[An attention unit solves MSD and LSD decision, Theorem C.2]
  \label{thm:attention-solves-msd-lsd}
  \uses{def:transformer, def:tf-solves, def:msd-dec, def:lsd-dec, def:attention-solves-predicate}
  An attention unit with input and output MLPs and parameters $d = \ell$,
  $d_{\mathrm{in}} = \ell + 1$, $d_{\mathrm{out}} = 1$, $m \ge \ell + 1$ can
  solve $\MSD_{n,\ell,t}$ and $\LSD_{n,\ell,t}$ for any $t \in [0,1]$; that is,
  $\AttentionSolves{\MSD_{n,\ell,t}}$ and $\AttentionSolves{\LSD_{n,\ell,t}}$.
\end{theorem}
\begin{proof}
  \uses{thm:attention-solves-maxip-minip, thm:attention-solves-ov, lem:mlp-normalize, lem:cosine-normalized-dot, def:msd-dec, def:lsd-dec, def:cosine}
  For $\MSD_{n,\ell,0}$ the answer is always yes for nonzero binary inputs (all
  cosine similarities are nonnegative), so a constant-output MLP suffices. For
  $\LSD_{n,\ell,0}$, the condition $\cossim{v_i}{v_j}\le 0$ is equivalent, for
  nonzero binary vectors, to $\inner{v_i}{v_j}=0$, so
  Theorem~\ref{thm:attention-solves-ov} solves this threshold-zero case.

  Now assume $t>0$. Given an $\MSD_{n,\ell,t}$ instance
  $v_1, \ldots, v_n$ with $V_{i,:} = v_i$, set
  $x_i = \left(\frac{v_i}{\norm{v_i}}, 1\right) \in \R^{\ell+1}$,
  $x_{n+1} = (0, \ldots, 0, t+1)$, and $X$ with rows $x_i$. Take the score
  bilinear form to be $M_{Q,K} = 3\log n \cdot I_{\ell+1}$ and
  $\mathcal V = (1, \ldots, 1, 0)$. By
  Lemma~\ref{lem:cosine-normalized-dot}, cosine similarity is exactly the inner
  product after Euclidean-normalizing the embeddings. Therefore the analysis of
  Theorem~\ref{thm:attention-solves-maxip-minip} (Max-IP case) applies once
  $\varphi_1$ produces the normalized rows $x_i$; the required normalizing MLP
  exists by Lemma~\ref{lem:mlp-normalize}, applied with separation threshold
  $\tau=t+1>1$ and $\rho=1$. The proof for $\LSD_{n,\ell,t}$ with $t>0$ is
  identical to the Min-IP case of Theorem~\ref{thm:attention-solves-maxip-minip}
  after applying the same normalizing $\varphi_1$
  (Lemma~\ref{lem:mlp-normalize} with $\tau=t+1$ and $\rho=1$).
\end{proof}

\chapter{MLP constructions (Appendix D)}

\begin{lemma}[Continuous thresholding MLP, Lemma D.1]
  \label{lem:mlp-threshold}
  \uses{lem:continuous-clipped-affine, lem:clipped-affine-values, lem:continuous-finite-product, lem:finite-product-zero-one}
  For any $a, b \in \R$ with $b > a$, there exists a continuous function
  $f : \R^{\ell} \to \R$ such that
  \[
    f(x) = \begin{cases}
      1 & \text{if } x[i] \ge b \text{ for some } 1 \le i \le \ell,\\
      0 & \text{if } x[i] < a \text{ for all } 1 \le i \le \ell.
    \end{cases}
  \]
  More formally, the existential witness consists of a function $f$, a proof
  that $f$ is continuous, a proof of
  $(\exists i,\ x[i]\ge b)\Rightarrow f(x)=1$, and a proof of
  $(\forall i,\ x[i]<a)\Rightarrow f(x)=0$. Proving only that some continuous
  function exists, for example the constant zero function, omits the behavioral
  guarantees and is not this lemma.
\end{lemma}
\begin{proof}
  \uses{lem:continuous-clipped-affine, lem:clipped-affine-values, lem:continuous-finite-product, lem:finite-product-zero-one, lem:continuity-closure}
  Define $g : \R \to \R$ by
  \[
    g(y)=\max\!\left(0,\min\!\left(1,\frac{y-a}{b-a}\right)\right).
  \]
  By Lemma~\ref{lem:continuous-clipped-affine}, $g$ is continuous, and by
  Lemma~\ref{lem:clipped-affine-values}, $g(y)=1$ when $y\ge b$ and $g(y)=0$
  when $y<a$. Let
  $f(x)=1-\prod_{i=1}^{\ell}(1-g(x[i]))$. Each coordinate map
  $x\mapsto 1-g(x[i])$ is continuous by composition with the coordinate
  projection and subtraction from a constant, so
  Lemma~\ref{lem:continuous-finite-product} gives continuity of
  $x\mapsto \prod_i(1-g(x[i]))$, and Lemma~\ref{lem:continuity-closure} gives
  continuity of $f$.

  If $x[i_0]\ge b$ for some $i_0$, then $g(x[i_0])=1$, so the $i_0$ factor
  $1-g(x[i_0])$ is $0$. Lemma~\ref{lem:finite-product-zero-one} gives that the
  whole product is $0$, hence $f(x)=1$. If $x[i]<a$ for all $i$, then every
  $g(x[i])=0$, so every factor is $1$; the all-one part of
  Lemma~\ref{lem:finite-product-zero-one} gives that the product is $1$, hence
  $f(x)=0$.

  In Lean, the continuity subproof should be organized as follows. First obtain
  `hg : Continuous g` from Lemma~\ref{lem:continuous-clipped-affine}. Then prove
  the product is continuous using `continuous_finsetProd Finset.univ`; for each
  coordinate $i$, the factor continuity is
  `continuous_const.sub (hg.comp (continuous_apply i))`. Finally close the
  continuity of $f$ by subtracting this product from `continuous_const`.
\end{proof}

\begin{lemma}[Continuous coordinate-appending MLP, Lemma D.2]
  \label{lem:mlp-append}
  \uses{lem:continuity-closure}
  For every $\tau>1$, there exists a continuous function
  $f : \R^{\ell} \to \R^{\ell+1}$ such that
  \[
    f(x) = \begin{cases}
      (x, 1) & \text{if } x[\ell] \le 1,\\
      (0, x) & \text{if } x[\ell] \ge \tau.
    \end{cases}
  \]
  The formal witness must include the continuity proof and both regional
  equalities:
  $x[\ell]\le1\Rightarrow f(x)=(x,1)$ and
  $x[\ell]\ge\tau\Rightarrow f(x)=(0,x)$. Equivalently, in coordinate form,
  for all $r\le\ell$:
  \[
    x[\ell]\le1 \Rightarrow
    \bigl(f(x)[r]=x[r]\ (r\le \ell)\ \wedge\ f(x)[\ell+1]=1\bigr),
  \]
  and
  \[
    x[\ell]\ge\tau \Rightarrow
    \bigl(f(x)[1]=0\ \wedge\ f(x)[r+1]=x[r]\ (1\le r\le\ell)\bigr).
  \]
  The proposition carried by this lemma is the conjunction
  \[
    \exists f,\ \operatorname{Continuous}(f)\ \wedge\
    \forall x,\ x[\ell]\le1\Rightarrow f(x)=(x,1)\ \wedge\
    \forall x,\ x[\ell]\ge\tau\Rightarrow f(x)=(0,x).
  \]
  The two implication conclusions are equalities of vectors, not `True'
  placeholders; a constant function cannot witness this lemma.
\end{lemma}
\begin{proof}
  \uses{lem:continuity-closure}
  Define $g : \R \to \R$ by $g(x) = 1$ if $x \le 1$,
  $g(x) = \frac{\tau-x}{\tau-1}$ if $1 < x < \tau$, and $g(x) = 0$ if
  $x \ge \tau$, which is continuous. Let $f_1(x) = (x, 1)$ and
  $f_2(x) = (0, x)$ in $\R^{\ell+1}$, both continuous. Set
  \[
    f(x) = g(x[\ell])\cdot f_1(x) + (1 - g(x[\ell]))\cdot f_2(x).
  \]
  By Lemma~\ref{lem:continuity-closure}, $f$ is continuous, and one checks
  directly that it equals $(x,1)$ when $x[\ell] \le 1$ and $(0,x)$ when
  $x[\ell] \ge \tau$ (and interpolates continuously in between).
\end{proof}

\begin{lemma}[Continuous normalizing MLP, Lemma D.3]
  \label{lem:mlp-normalize}
  \uses{lem:continuity-closure, lem:mlp-append}
  For every $\tau>1$ and $\rho>0$, there exists a continuous function
  $f : \R^{\ell} \to \R^{\ell+1}$ such that, for every $x$ with
  $\norm{x}\ge \rho$,
  \[
    f(x) = \begin{cases}
      \left(\frac{x}{\norm{x}}, 1\right) & \text{if } x[\ell] \le 1,\\
      \left(0, \frac{x}{\norm{x}}\right) & \text{if } x[\ell] \ge \tau.
    \end{cases}
  \]
  The formal witness must include continuity on all of $\R^\ell$ and the two
  displayed normalization equalities on the restricted domain $\norm{x}\ge\rho$.
  It must therefore construct a safe denominator or equivalent continuous
  extension. In full, the theorem asserts
  \[
    \exists f,\ \operatorname{Continuous}(f)\ \wedge\
    \forall x,\ \norm{x}\ge\rho \Rightarrow
      \left(x[\ell]\le1\Rightarrow
        f(x)=\left(\frac{x}{\norm{x}},1\right)\right)
      \wedge
      \left(x[\ell]\ge\tau\Rightarrow
        f(x)=\left(0,\frac{x}{\norm{x}}\right)\right).
  \]
  The norm lower bound, the two guard conditions, and the two vector equalities
  are all part of the statement. A proof that returns an arbitrary continuous
  function and proves only `True' under the norm hypothesis does not satisfy the
  blueprint node.
\end{lemma}
\begin{proof}
  \uses{lem:mlp-append, lem:continuity-closure, def:l2norm}
  Let $g$ be the same continuous cutoff as in Lemma~\ref{lem:mlp-append}. Define
  the safe denominator $s_\rho(x)=\max(\norm{x},\rho/2)$, which is continuous and
  strictly positive. Let
  $f_1(x)=\left(\frac{x}{s_\rho(x)},1\right)$ and
  $f_2(x)=\left(0,\frac{x}{s_\rho(x)}\right)$, both continuous on all of
  $\R^\ell$. Then
  \[
    f(x) = g(x[\ell])\cdot f_1(x) + (1 - g(x[\ell]))\cdot f_2(x)
  \]
  is continuous by Lemma~\ref{lem:continuity-closure}. On inputs with
  $\norm{x}\ge\rho$, the safe denominator equals $\norm{x}$; if in addition
  $x[\ell]\le 1$ then $g(x[\ell])=1$, and if $x[\ell]\ge\tau$ then
  $g(x[\ell])=0$. Hence the required piecewise values hold on the separated
  nonzero input region used by the construction.
\end{proof}
